## Combining Data

:::{.callout-note title="Learning Goals" collapse="false"}
After this lesson, you should be able to:

* Explain what a relational dataset is
* Use the dplyr package to join data based on common columns
* Describe the different types of joins
* Identify which types of joins to use when faced with a relational dataset
:::


## Relational Datasets

Think about how you would organize restaurant reviews data. Each restaurant has
a name, address, phone number, operating hours, and other restaurant level
details. Because the number of reviews for each restaurant will vary, we could
put the data in a single table with one row for each review. Then the table is
at the review level, and we'd have to repeat restaurant level details across
all reviews for each restaurant.

The review level table is convenient if we want to analyze reviews, but not so
convenient if we want to analyze restaurants. For instance, suppose we want to
count how many restaurants open before 10 a.m. In order to avoid counting each
restaurant multiple times, we have to reduce the table to one row per
restaurant before we compute the count.

The problem with putting the restaurant review data in a single table is that
it consists of observations at two different levels: the restaurant level and
the review level. From this perspective, an intuitive solution is to put the
data in two tables: a "restaurants" table where each row is a restaurant and a
"reviews" table where each row is a review. Then we can choose the most
suitable table for each question we want to answer.

Each review is associated with a restaurant, and each restaurant is associated
with some reviews, so the two tables are related even though they're separate.
We can keep track of the relationship by including a column for restaurant name
(or a unique restaurant identifier) in both tables. This makes the data
**relational**: multiple tables where the relationships between them are
expressed through columns in common.

:::{.callout-note}
Most database software are designed to efficiently store and query relational
data, so in computing contexts, **database** is often synonymous with
relational data.
:::

For some questions, we'll need to use both the restaurants table and the
reviews table. For example, suppose we want to count how many restaurants open
before 10 a.m. *and* have at least 1 five-star review. We can use the reviews
table to compute the number of five-star reviews for each restaurant, combine
these with the restaurants table, and then count restaurants that meet our
conditions in the resulting table. Operations that combine two related tables
based on columns they have in common are called **joins**. The two tables are
conventionally called the **left table** and **right table**.

There are a few different kinds of joins. We'll use a simplified, fictitious
version of the restaurant reviews data to demonstrate some of them. The
`restaurants` table contains names and phone numbers for three restaurants:

```{r}
restaurants = data.frame(
  id = c(1, 2, 3),
  name = c("Alice's Restaurant", "The Original Beef", "The Pie Hole"),
  phone = c("555-3213", "555-1111", "555-9983")
)
restaurants
```

The `reviews` table contains scores from five restaurant reviews:

```{r}
reviews = data.frame(
    id = c(4, 2, 1, 2, 2),
    score = c(4.2, 3.5, 4.7, 4.8, 4.0)
)
reviews
```


## The dplyr Package

The [dplyr][] package provides a wide variety of functions for working with
data frames. In this chapter, we'll focus almost exclusively on the functions
for combining data frames, but the other parts of the package are also useful.
Like most Tidyverse packages, it comes with detailed [documentation][dplyr] and
a [cheatsheet][dplyr-cheat].

[dplyr]: https://dplyr.tidyverse.org/
[dplyr-cheat]: https://github.com/rstudio/cheatsheets/blob/main/data-transformation.pdf

Install the package if you haven't already, and then load it:

```{r}
# install.packages("dplyr")
library("dplyr")
```

:::{.callout-note}
If you've ever used SQL, you're probably familiar with relational datasets and
recognize commands like `SELECT`, `LEFT JOIN`, and `GROUP BY`. The dplyr
package borrows some of the ideas of SQL, and provides `select`, `left_join`,
and `group_by` functions (among others) that are reminiscent of their SQL
counterparts.

If you haven't used SQL, don't worry---all of the functions will be explained
in detail.
:::


## Inner Joins

An **inner join** only keeps rows from the left table if they match rows in the
right table and vice-versa.

You can use dplyr's `inner_join` function to join two data frames with an inner
join. The first argument is the left table and the second argument is the right
table. Set the `by` parameter to the name of the column to use to match the
rows. Try joining the `restaurants` table and the `reviews` table:

```{r}
inner_join(restaurants, reviews)
```

The inner join keeps rows where `id` is `1` or `2`, since these values appear
in both tables. It drops the row where `id` is `3` in `restaurants` and the row
where `id` is `4` in `reviews`. The resulting table has 4 rows and the columns
from both data frames.

:::{.callout-tip}
If the columns you want to use to join two data frames have different names,
set `by = join_by(LCOL == RCOL)`, replacing `LCOL` with the name of the column
in the left table and `RCOL` with the name of the column in the right table.
:::


## Left & Right Joins

A **left join** keeps all rows from the left table and only keeps rows from the
right table if they match. Missing values fill any spaces where there was no
match.

You can use dplyr's `left_join` function to join two data frames with a left
join. The arguments are the same as for the `inner_join` function. Try a left
join on the restaurant reviews data:

```{r}
left_join(restaurants, reviews)
```

The left join keeps all of the rows from `restaurants`, and matches rows from
`reviews` when possible. There are no reviews where the `id` is `3`, so `score`
is missing for that row.

A left join is asymmetric, so switching the order of the tables will generally
produce a different result:

```{r}
left_join(reviews, restaurants)
```

A **right join** is equivalent to a left join with the order of the tables
switched. Because of this, some relational data tools don't have a right join
command (only a left join command). You can use dplyr's `right_join` function
to join two data frames with a right join.


## Full Joins

A **full join** keeps all rows from both tables. Missing values fill any spaces
where there was no match.

You can use dplyr's `full_join` function to join two data frames with a full
join. The arguments are the same as for the `inner_join` and `left_join`
function. Try a full join on the restaurant reviews data:

```{r}
full_join(restaurants, reviews)
```

:::{.callout-note title="See Also"}
There are a few more kinds of joins. See the dplyr [documentation][dplyr] for
details.
:::


## Case Study: CA Crash Reporting System

The California Highway Patrol publish data about vehicle crashes in the state
as the California Crash Reporting System (CCRS). The CCRS is a relational data
set with three tables per year which describe crash events, parties involved,
and all injuries, witnesses, and passengers. Let's use the 2024 CCRS data for
Sacramento and Yolo Counties to compute statistics about crashes in those
counties.

:::{.callout-important}
[Click here][crash-data] to download the 2024 Sacramento & Yolo Crash dataset
(3 CSV files).

[crash-data]: https://ucdavis.box.com/s/kjxowylvg3cfqgnofh03gn5e3xicsobt

If you haven't already, we recommend you create a directory for this workshop.
In your workshop directory, create a `data/` subdirectory. Download and save
the dataset in the `data/` subdirectory.
:::

:::{.callout-note title="Documentation for the 2024 Sacramento & Yolo Crash Dataset" collapse=true}

The dataset consists of three tables:

* `2024_sac-yolo_crashes.csv`, where each row is a crash.
* `2024_sac-yolo_parties.csv`, where each row is a person directly involved in
  the crash (not a witness or passenger).
* `2024_sac-yolo_injured-witness-passenger.csv`, where each row is an injured
  person (including drivers), witness, or passenger.

[Click here][ccrs-docs] to download the documentation for the source dataset.

[ccrs-docs]: https://ucdavis.box.com/s/pfgbup6fw17nq7vajo1e05eoj0w57ygi

This dataset is a subset of the much larger [CA Crash Reporting System dataset][ccrs].

[ccrs]: https://data.ca.gov/dataset/ccrs
:::

To get started, let's read each of the three tables in this dataset into R.
We'll use the [readr][] package (a part of the Tidyverse), since its functions
are more efficient and robust than R's built-in data reader functions.
Specifically, we'll use the `read_csv` function to read the files.

[readr]: https://readr.tidyverse.org/

In the crashes table, the `SketchDesc` column contains many blanks, which
inteferes with the `read_csv` function's ability to infer the column type. To
fix this, we'll set the column type manually. In the injured-witness-passenger
table, the `SeatPositionOther` column has the same problem, and we'll use the
same fix. Here's the code to read the three files:

```{r}
library("readr")

crashes = read_csv(
  "data/ca_crash_reporting/2024_sac-yolo_crashes.csv",
  col_types = list(SketchDesc = "c")
)

parties = read_csv("data/ca_crash_reporting/2024_sac-yolo_parties.csv")

injured = read_csv(
  "data/ca_crash_reporting/2024_sac-yolo_injured-witness-passenger.csv",
  col_types = list(SeatPositionOther = "c")
)
```

Each table has many columns we could use. To stay focused, we'll try to answer
two questions:

1. How do lighting conditions interact with driver age?
2. Among injured people, do drivers or passengers experience more severe
   injuries? Does it matter whether they're in the vehicle of the driver at
   fault?

There are, of course, many other ways to explore this dataset.


### Question 1

In this section, we'll try to address this question:

> How do lighting conditions interact with driver age?

Let's work on driver age first. In the parties table, the `StatedAge` column
gives the age of each party involved in the crash. The parties table includes
drivers, but also pedestrians, cyclists, and other parties; this is indicated
by the `PartyType` column. We only want drivers, so we'll filter the rows:

```{r}
library("dplyr")

drivers = filter(parties, PartyType == "Driver")
```

This reduces the table from `r nrow(parties)` to `r nrow(drivers)` rows.

Now we need to get the lighting conditions associated with each crash. The
`LightingCode` and `LightingDescription` columns in the crashes table provide
this information. The two columns are just different ways of describing the
same thing, which we can verify by counting all combinations of them:

```{r}
count(crashes, LightingCode, LightingDescription)
```

We'll keep both, since the codes in `LightingCode` are easier to type (for
instance, for filtering) but the descriptions in `LightingDescription` are
easier to understand.

We can attach the lighting conditions to the drivers data frame with a join.
We want to get a lighting condition for every driver, so we'll use a left join
with the drivers data frame as the left table. We can use the `CollisionId`
column, which is shared by both data frames, as the key for the join:

```{r}
joined = left_join(
  drivers,
  # We only care about 3 columns in the crashes table.
  crashes[c("CollisionId", "LightingCode", "LightingDescription")],
  by = "CollisionId"
)
```

Notice that the resulting data frame has the same number of rows as the drivers
data frame:

```{r}
nrow(drivers) == nrow(joined)
```

This indicates that we didn't duplicate any drivers in the join.

Driver age is a continous feature, while lighting condition in categorical. So
we can use a density plot or histogram to show age, with different colors to
show lighting condition. For instance, here's a density plot:

```{r}
library("ggplot2")

ggplot(joined) +
  aes(x = StatedAge, color = LightingDescription) +
  geom_density() +
  # Increase the number of x-axis tick marks.
  scale_x_continuous(n.breaks = 20) +
  # Position legend at the bottom and decrease font size (since it's wide).
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 6),
    legend.title = element_text(size = 6)
  )
```

Alternatively, here's a histogram:

```{r}
ggplot(joined) +
  aes(x = StatedAge, fill = LightingDescription) +
  geom_histogram(color = "black") +
  scale_x_continuous(n.breaks = 20) +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 6),
    legend.title = element_text(size = 6)
  )
```

In both cases, the number of lighting categories makes the plots difficult to
read. Let's create a new feature to separate out the dark categories from the
others. The dark categories have `LightingCode` equal to `C`, `D`, and `E`:

```{r}
joined$is_dark = joined$LightingCode %in% c("C", "D", "E")
```

We can use the new `is_dark` feature to vary the line type in the density plot:

```{r}
ggplot(joined) +
  aes(x = StatedAge, color = LightingDescription, linetype = is_dark) +
  geom_density() +
  scale_x_continuous(n.breaks = 20) +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 6),
    legend.title = element_text(size = 6)
  )
```

Or to facet the histogram:

```{r}
ggplot(joined) +
  aes(x = StatedAge, fill = LightingDescription) +
  geom_histogram(color = "black") +
  scale_x_continuous(n.breaks = 10) +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 6),
    legend.title = element_text(size = 6)
  ) +
  facet_wrap(~ is_dark)
```

Based on the plots, we can see that most crashes happen in daylight and involve
younger drivers (under 50). In daylight crashes, drivers from age 20-40 are
almost equally likely, whereas in dark crashes there are sharp peaks around 25.
The "Dark - Street Lights Not Functioning" category is notably different from
the others, with a peak around age 32, but the histogram shows that this
category as also very small, so the peak is likely just a quirk of this sample
rather than a general tendency in the larger population (and it would be quite
surprising if older drivers were more likely to encounter non-functional street
lights).


### Question 2

Now we'll try to address the second question:

> Among injured people, do drivers or passengers experience more severe
> injuries? Does it matter whether they're in the vehicle of the driver at
> fault?

For the first part of the question, we need to look at the injured table. The
table includes an `InjuredPersonType` column that indicates which people are
passengers and which are drivers. A complication here is that the
`IsPassengerOnly` column also appears to indicate passengers, and the doesn't
match up with `InjuredPersonType`:

```{r}
count(
  injured,
  PassengerType = InjuredPersonType == "Passenger",
  IsPassengerOnly
)
```

We can compare `IsPassengerOnly` with `ExtentOfInjuryCode` to see that the
column seems to indicate *uninjured passengers*:

```{r}
count(injured, IsPassengerOnly, ExtentOfInjuryCode)
```

So let's filter the table to keep only the rows where the person type is
`Driver` or `Passenger`:

```{r}
injured_in_car = filter(
  injured,
  InjuredPersonType %in% c("Driver", "Passenger")
)
```

The nature of injure is in the `ExtentOfInjuryCode` column. To make our
visualizations easier to read, let's put the categories in order from least to
most severe (with the understanding that some categories are vague):

```{r}
injured_in_car$ExtentOfInjuryCode = factor(
  injured_in_car$ExtentOfInjuryCode,
  c("OtherVisibleInactive", "ComplaintOfPainInactive", "PossibleInjury",
    "SuspectMinor", "SuspectSerious", "Fatal")
)
```

Now we can make a bar plot to address the first part of the question:

```{r}
ggplot(injured_in_car) +
  aes(x = ExtentOfInjuryCode, fill = InjuredPersonType) +
  geom_bar(position = "dodge") +
  theme(
    legend.position = "bottom",
    # Rotate axis labels to 45 degrees and align at right edge.
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

The plot suggests that far more drivers are injured in crashes than passengers.
That said, it's important to keep in mind that the plot shows all crashes. If
we filter the dataset to only the crashes where passengers were present, the
result might be very different. Try visualizing this as an exercise!

Let's address the second part of the question about being in the vehicle of the
driver at fault. The party at fault is indicated by the `IsAtFault` column in
the parties table. We need to join this to the `injured_in_car` data frame.
Since collisions usually involve multiple parties, we need to use both
`CollisionId` and `PartyNumber` as keys in the join. We'll use a left join with
`injured_in_car` as the left table:

```{r}
joined = left_join(
  injured_in_car,
  parties[c("CollisionId", "PartyNumber", "IsAtFault")],
  by = c("CollisionId", "PartyNumber")
)
```

We can use the joined `IsAtFault` column to facet the bar plot we made earlier:

```{r}
ggplot(joined) +
  aes(x = ExtentOfInjuryCode, fill = InjuredPersonType) +
  geom_bar(position = "dodge") +
  facet_wrap(~ IsAtFault) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

From the plot, we can see that fewer injured people were in at fault vehicles
overall, but there were still more drivers injured than passengers. This could
be due to crashes where only one driver is at fault, but more than two vehicles
are involved. Once again, we could check this through further inspection and
visualization of the dataset. Try writing the code to do so!
