# String & Date Processing {#sec-string-date-processing}

:::{.callout-note title="Learning Goals" collapse="false"}
After this lesson, you should be able to:

* Explain why we use special data structures for dates & times
* Identify the correct data structure for a given date/time
* Use the lubridate package to parse dates
* Use the date format string mini-language
* Use escape codes in strings to represent non-keyboard characters
* Explain what a text encoding is
* Use the stringr package to detect, extract, and change patterns in strings
* Use the regular expressions mini-language
:::

This chapter is part 1 (of 2) of _Cleaning & Reshaping Data_, a workshop series
about how to prepare data for analysis. The major topics of this chapter are
how to convert dates and times into appropriate R data types and how to extract
and clean up data from strings (including numbers with non-numeric characters
such as `$`, `%`, and `,`).


## The Tidyverse {#sec-the-tidyverse}

For working with dates, times, and strings, we recommend using packages from
the [Tidyverse][], a popular collection of packages for doing data science.
Compared to R's built-in functions, we've found that the functions in Tidyverse
packages are generally easier to learn and use. They also provide additional
features and have more robust support for characters outside of the Latin
alphabet.

Although they're developed by many different members of the R community,
Tidyverse packages follow a unified design philosophy, and thus have many
interfaces and data structures in common. The packages provide convenient and
efficient alternatives to built-in R functions for many tasks, including:

[Tidyverse]: https://www.tidyverse.org/

* Reading and writing files (package readr)
* Processing dates and times (packages lubridate, hms)
* Processing strings (package stringr)
* Reshaping data (package tidyr)
* Making visualizations (package ggplot2)
* And more

Think of the Tidyverse as a different dialect of R. Sometimes the syntax is
different, and sometimes ideas are easier or harder to express concisely. As a
consequence, the Tidyverse is sometimes polarizing in the R community. It's
useful to be literate in both base R and the Tidyverse, since both are popular.

One major advantage of the Tidyverse is that the packages are usually
well-documented and provide lots of examples. Every package has a documentation
website and the most popular ones also have [cheatsheets][].

[cheatsheets]: https://posit.co/resources/cheatsheets/


## Dates & Times

Given a dataset with dates or times, you might want to:

* Sort the observations chronologically
* Compute the times between observations
* Add or subtract a time offset
* Get just the month, year, hour, etc. of each observation

You can do all of these things and more with R, but only if your dates and
times are represented by appropriate data types.


### The lubridate Package {#sec-the-lubridate-package}

As explained in @sec-the-tidyverse, we recommend the Tidyverse packages for
working with dates and times over other packages or R's built-in functions.
There are two:

* [lubridate][], the primary package for working with dates and times
* [hms][], a package specifically for working with time durations

This chapter only covers lubridate, since it's more useful in most situations.
The package has detailed [documentation][lubridate] and a
[cheatsheet][lubridate-cheatsheet].

[lubridate]: https://lubridate.tidyverse.org/
[lubridate-cheatsheet]: https://rawgit.com/rstudio/cheatsheets/main/lubridate.pdf
[hms]: https://hms.tidyverse.org/

:::{.callout-note}
A relatively new package, [clock][], tries to solve some problems with the
`Date` class people have identified over the years. The package is in the r-lib
collection of packages, which provide low-level functionality complementary to
the Tidyverse. Eventually, it may be preferable to use the classes in clock
rather than the `Date` class, but for now, the `Date` class is
still suitable for most tasks.

[clock]: https://clock.r-lib.org/
:::

You'll have to install the package if you haven't already, and then load it:

```{r}
# install.packages("lubridate")
library("lubridate")
```

### Parsing Dates {#sec-parsing-dates}

Many popular file formats for data, such as CSV, *do not* store metadata about
which values are dates and times, so these values get typed as strings. Dates
and times extracted from text are also naturally strings.

The first step to working with dates and times typed as strings is to **parse**
them: break them down into their components and cast them to more suitable data
types. To demonstrate parsing dates, consider a vector of date strings:

```{r}
date_strings = c("Jan 10, 2021", "Sep 3, 2018", "Feb 28, 1982")
date_strings
```

*You* can tell that these are dates, but as far as R is concerned, they're
text.

The lubridate package provides a variety of functions to automatically parse
strings into date or time objects that R understands. These functions are named
with one letter per component of the date or time. The order of the letters
must match the order of the components in the string you want to parse.

The example strings have the month (`m`), then the day (`d`), and then the year
(`y`), so you can use the `mdy` function to parse them automatically:

```{r}
dates = mdy(date_strings)
dates
class(dates)
```

Notice that R prints the dates differently now. Thanks to the `mdy` function,
they now have class `Date`, one of R's built-in classes for representing dates.
R recognizes that the dates are in fact dates, so they're ready to use in an
analysis.

There is [a complete list of the automatic parsing functions][lubridate-parse]
in the lubridate documentation.

[lubridate-parse]: https://lubridate.tidyverse.org/reference/index.html#date-time-parsing

Occasionally, a date or time string may have a format that lubridate can't
parse automatically. In that case, you can use the `fast_strptime` function to
describe the format in detail.

At a minimum, the function requires two arguments: a vector of strings to parse
and a **format string**. The format string describes the format of the dates.
In a format string, a percent sign `%` followed by a character is called a
specifier and represents a component of a data or time. Here are some useful
specifiers:

| Specification | Description      | 2015-01-29 21:32:55
|:------------- |:---------------- |:-------------------
| `%Y`          | 4-digit year     | 2015
| `%m`          | 2-digit month    | 01
| `%d`          | 2-digit day      | 29
| `%H`          | 2-digit hour     | 21
| `%M`          | 2-digit minute   | 32
| `%S`          | 2-digit second   | 55
| `%%`          | literal %        | %
| `%y`          | 2-digit year     | 15
| `%B`          | full month name  | January
| `%b`          | short month name | Jan


Other characters in the format string don't have any special meaning. Write the
format string so that it matches the format of the dates you want to parse.

:::{.callout-note}
Most programming languages use format strings to parse and format dates and
times. There isn't a standard, but the idea seems to have originated with the
`strptime` and `strftime` functions in the C programming language.

You can find a complete list of specifiers for lubridate in `?fast_strptime`.
:::

To demonstrate `fast_strptime`, let's try parsing an unusual time format:

```{r}
time_string = "6 minutes, 32 seconds after 10 o'clock"
time = fast_strptime(time_string, "%M minutes, %S seconds after %H o'clock")
time
class(time)
```

R represents datetimes (which combine a date and a time) with the classes
`POSIXlt` and `POSIXct`. There's no built-in class to represent times alone,
which is why the result in the example above includes a date.

:::{.callout-note}
Internally, a `POSIXlt` object is a list with elements to store different date
and time components. On the other hand, a `POSIXct` object is a single floating
point number (type `double`).

If you want to store your time data in a data frame, use `POSIXct` objects,
since data frames don't work well with columns of lists.

You can control whether `fast_strptime` returns a `POSIXlt` or `POSIXct` object
by setting the `lt` parameter to `TRUE` or `FALSE`:

```{r}
time_ct = fast_strptime(time_string, "%M minutes, %S seconds after %H o'clock",
  lt = FALSE)

class(time_ct)
```
:::


### Creating Dates

Another common task is combining the numeric components of a date or time into
a single object. You can use the `make_date` and `make_datetime` functions to
do this. The parameters are named for the different components. For example:

```{r}
events = make_datetime(
  year = c(2023, 2002),
  month = c(1, 8),
  day = c(10, 16),
  hour = c(8, 14),
  min = c(3, 59)
)
events
```

These functions are vectorized, so you can use them to combine the components
of many dates or times at once. They're especially useful for reconstructing
dates and times from tabular datasets where each component is stored in a
separate column.


### Extracting Components

Once you have dates or times, you might want to extract components from them.
For instance, suppose we want to get the years from the dates we parsed
earlier. You can use lubridate functions to get or set the components. These
functions usually have the same name as the component. For instance, the `year`
function gets the year:

```{r}
year(dates)
```

Similarly, the `month` function gets the month:

```{r}
month(dates)
```

See [the lubridate documentation][lubridate] for even more details about
what you can do.


### Durations & Periods

Occasionally, you might need to adjust dates or times by adding an offset. For
example, suppose we want to add 30 days to the dates from earlier. The
`duration` function creates a **duration** (class `Duration`), which represents
a fixed amount of time. So to add 30 days:

```{r}
dates + duration(30, "days")
```

What if we want to add 1 month to the dates instead of 30 days? The length of a
month varies. The `duration` function gives the average amount of time in a
month, but that won't be correct for some months. The solution is to use a
**period** (class `Period`) rather than a duration. The `period` function
creates a period. So to add 1 month (of variable length):

```{r}
dates + period(1, "month")
```

This increments each date by exactly 1 month, regardless of how many days that
is.

:::{.callout-tip}
There are helper functions for common periods named after the associated unit
of time. For example, another way add a 1-month period to the dates is with the
`months` function:

```{r}
dates + months(1)
```

Similarly, there are helper functions for common durations. These are also
named after the associated unit of time, but always begin with the prefix `d`.
So to add a 30-day duration to the dates:

```{r}
dates + ddays(30)
```
:::


## Case Study: CA Parks & Recreation Fleet

The government of California publishes data about its fleet of vehicles on the
[California Open Data portal][data.ca.gov]. As of March 2025, the dataset
includes all non-confidential vehicles owned by agencies from 2015-2023. We'll
use a subset of this data to compute how many vehicles the CA Department of
Parks and Recreation purchased each year from 2019 to 2023. The dataset is
published as a messy CSV, so we'll need to do some cleaning and parse the dates
in order to use it.

[data.ca.gov]: https://data.ca.gov/

:::{.callout-important}
[Click here][ca-parks-fleet] to download the CA Parks & Recreation Fleet
dataset.

[ca-parks-fleet]: https://ucdavis.box.com/s/6j40xd44miz7x6qlvgga9eieu7ya6xo3

If you haven't already, we recommend you create a directory for this workshop.
In your workshop directory, create a `data/` subdirectory. Download and save
the dataset in the `data/` subdirectory.
:::

:::{.callout-note title="Documentation for the CA Parks & Recreation Fleet Dataset" collapse="true"}
Each row in the dataset contains measurements from one vehicle-year
combination.

[Click here][ca-fleet-docs] to download the documentation for the columns.

[ca-fleet-docs]: https://ucdavis.box.com/s/v6agya4830tz1lkgq8toej5q1b9j5ntf

This dataset is a subset of the much larger [CA State Fleet dataset][ca-fleet].

[ca-fleet]: https://data.ca.gov/dataset/california-state-fleet
:::

To get started, read the data set from wherever you saved it:

```{r}
fleet = read.csv("data/2015-2023_ca_parks_fleet.csv")
head(fleet)
```

Since we want to compute the number of acquisitions each year, we'll focus on
two columns: `acquisition_method` and `acquisition_delivery_date`. The
`acquisition_method` column indicates whether the vehicle was purchased,
donated, transferred from a different agency, or something else. We're only
interested in purchases, so we can filter the others out:

```{r}
library("dplyr")

fleet = filter(fleet, acquisition_method == "Purchase")
```

In order to determine how many vehicles were purchased each year, we need to
extract the year from `acquisition_delivery_date`. Let's take a look at this
column:

```{r}
head(fleet$acquisition_delivery_date)
```

It looks like the dates are in month-day-year format, so we can use lubridate's
`mdy` function:

```{r}
dates = mdy(fleet$acquisition_delivery_date)
```

Uh-oh. According to the warning, lubridate couldn't parse some of the dates.
This means they probably don't follow the month-day-year format. When lubridate
can't parse a date, it returns a missing value. We can use this to figure out
what the dates look like:

```{r}
is_bad_date = is.na(dates)
bad_dates = fleet$acquisition_delivery_date[is_bad_date]
head(bad_dates, 20)
```

Instead of month-day-year dates, they're numbers like `44832`, `44462`, and
`43451`. These numbers probably seem inscrutable, but there is an explanation
for them: Microsoft Excel, perhaps the most popular tool for working with
tabular data, stores dates by counting days from 31 December 1899. So 1 is 1
January 1900, 32 is 1 February 1900, and so on. Unfortunately, the Excel
developers incorrectly assumed 1900 was a leap year, so all of the counts were
off by 1 after 28 February 1900. Most modern Excel-compatible spreadsheet
programs fix this by counting from 30 December 1899 rather than 31 December, so
that only dates before 28 February have different numbers.

We can convert the numbers in the `bad_dates` variable into dates by treating
them as offsets to 30 December 1899:

```{r}
start_date = make_date(1899, 12, 30)

day_offsets = as.numeric(bad_dates)
fixed_dates = start_date + days(day_offsets)
head(fixed_dates, 20)
```

We can insert these back into the `dates` vector at their original positions.
Then we can replace the `acquisition_delivery_date` column with `dates`:

```{r}
dates[is_bad_date] = fixed_dates

fleet$acquisition_delivery_date = dates
```

With the dates in hand, we can compute the number of purchases from 2019 to
2023. Since each row in the dataset represents a vehicle-year, and we just want
to count vehicles, we must first get a subset of unique vehicles. We can do
this by grouping on the `equipment_number` column, which is a unique identifier
for each vehicle, and getting the first row in each group:

```{r}
by_vehicle = group_by(fleet, equipment_number)
vehicles = filter(by_vehicle, row_number() == 1)
```

We can compute the number of purchases by extracting the years with lubridate's
`year` function and then counting them with the `table` function:

```{r}
table(year(vehicles$acquisition_delivery_date))
```

It's unclear whether the dataset contains complete records for 2023, so the
count for that year might be too small. It's also likely that the agency
purchases vehicles months or years in advance of their delivery, so these
counts are more accurately described as deliveries rather than purchases.


## Case Study: Ocean Temperatures

The U.S. National Oceanic and Atmospheric Administration (NOAA) publishes ocean
temperature data collected by sensor buoys off the coast on the [National Data
Buoy Center (NDBC) website][ndbc]. California also has many sensors collecting
ocean temperature data that are not administered by the federal government.
Data from these is published on the [California Ocean Observing Systems
(CALOOS) Data portal][caloos].

[ndbc]: https://www.ndbc.noaa.gov/
[caloos]: https://data.caloos.org/

Suppose you're a researcher who wants to combine ocean temperature data from
both sources to use in R. Both publish the data in comma-separated value (CSV)
format, but record dates, times, and temperatures differently. Thus you need to
be careful that the dates and times are parsed correctly.

:::{.callout-important}
Download these two 2021 datasets:

1. `2021_noaa-ndbc_46013.txt`, from NOAA buoy 46013, off the coast of Bodega
   Bay ([DOWNLOAD][noaa-data])([source][noaa-source])
2. `2021_ucdavis_bml_wts.csv`, from the UC Davis Bodega Bay Marine Laboratory's
   sensors ([DOWNLOAD][bml-data])([source][bml-source])

[noaa-data]: https://drive.google.com/file/d/1w7WwX8yxs_1pQUSQreIE8v97ETeHKA__/view?usp=share_link
[noaa-source]: https://www.ndbc.noaa.gov/station_page.php?station=46013
[bml-data]: https://drive.google.com/file/d/1YZbVu30usB988O4idES-VqN3e8_a9OXM/view?usp=share_link
[bml-source]: https://data.caloos.org/#metadata/19946/station/data
:::

The NOAA data has a fixed-width format, which means each column has a fixed
width in characters over all rows. The [readr][] package provides a function
`read_fwf` that can automatically guess the column widths and read the data
into a data frame. The column names appear in the first row and column units
appear in the second row, so read those rows separately:

[readr]: https://readr.tidyverse.org/

```{r}
# install.packages("readr")
library("readr")

noaa_path = "data/ocean_data/2021_noaa-ndbc_46013.txt"
noaa_headers = read_fwf(noaa_path, n_max = 2, guess_max = 1)
noaa = read_fwf(noaa_path, skip = 2)
names(noaa) = as.character(noaa_headers[1, ])
names(noaa)[1] = "YY"
```

The dates and times for the observations are separated into component columns,
and the `read_fwf` function does not convert some of these to numbers
automatically. You can use `as.numeric` to convert them to numbers:

```{r}
cols = 2:5
noaa[cols] = lapply(noaa[cols], as.numeric)
```

Finally, use the `make_datetime` function to combine the components into
date-time objects:

```{r}
noaa_dt = make_datetime(year = noaa$YY, month = noaa$MM, day = noaa$DD,
  hour = noaa$hh, min = noaa$mm)
noaa$date = noaa_dt
head(noaa_dt)
```

That takes care of the dates in the NOAA data.

The Bodega Marine Lab data is CSV format, which you can read with `read.csv` or
the readr package's `read_csv` function. The latter is faster and usually
better at guessing column types. The column names appear in the first row and
the column units appear in the second row. The `read_csv` function handles the
names automatically, but you'll have to remove the unit row as a separate step:

```{r}
bml = read_csv("data/ocean_data/2021_ucdavis_bml_wts.csv")
bml = bml[-1, ]
```

The dates and times of the observations were loaded as strings. You can use
lubridate's `ymd_hms` function to automatically parse them:

```{r}
bml_dt = ymd_hms(bml$time)
bml$date = bml_dt
head(bml_dt)
```

Now you have date and time objects for both datasets, so you can combine the
two. For example, you could extract the `date` and water temperature columns
from each, create a new column identifying the data source, and then row-bind
the datasets together.


## Strings & Regular Expressions

Strings represent text, but even if your datasets are composed entirely of
numbers, you'll need to know how to work with strings. Text formats for data
are widespread: comma-separated values (CSV), tab-separated values (TSV),
JavaScript object notation (JSON), various markup languages (HTML, XML, YAML,
TOML), and more. When you read data in these formats into R, sometimes R will
correctly convert the values to appropriate non-string types. The rest of the
time, you need to know how to work with strings so that you can fix whatever
went wrong and convert the data yourself.

String processing encompasses a variety of tasks such as searching for patterns
within strings, extracting data from within strings, splitting strings into
component parts, and removing or replacing unwanted characters (excess
whitespace, punctuation, and so on).


### String Fundamentals

This section introduces several fundamental concepts related to working with
strings.


#### Escape Sequences

In a string, an **escape sequence** or **escape code** consists of a backslash
followed by one or more characters. Escape sequences make it possible to:

1. Write quotes or backslashes within a string
2. Write characters that don't appear on your keyboard (for example, characters
   in a foreign language)

For example, the escape sequence `\n` corresponds to the newline character.
Notice that the `message` function prints `\n` as a literal new line:

```{r}
x = "Hello\nNick"

message(x)
```

On the other hand, when you let R print a string or other object automatically
(without explicitly calling `message`), it prints a programmer-friendly
representation. For strings, the representation shows escape codes:

```{r}
x
```

This makes it easy to identify characters that are not normally visible, such
as whitespace, and also makes it easy to copy the string. You can use the
built-in `print` function to get a representation for any object.

:::{.callout-note}
To print the actual characters in a string, use the `message` function.

To print a representation of the characters in a string, use the `print`
function. The representation is useful to identify characters that are not
normally visible, such as tabs and the characters that mark the end of a line.

There are at least two more functions in R that can print things: `cat` and
`show`. The `cat` function is similar to `message`, but bypasses R's system for
keeping track of messages. The `show` function is similar to `print`, but
specifically designed for R's S4 objects, which we won't cover. You generally
shouldn't use these two functions directly.
:::

As another example, suppose we want to put literal quotation marks in a string.
We can either enclose the string in the other kind of quotation marks (single
versus double), or escape the quotation marks in the string:

```{r}
x = 'She said, "Hi"'
message(x)

y = "She said, \"Hi\""
message(y)
```

Since escape sequences begin with backslash, we also need to use an escape
sequence to write a literal backslash. The escape sequence for a literal
backslash is two backslashes:

```{r}
x = "\\"
message(x)
```

There's a complete list of escape sequences for R in the `?Quotes` help file.
Other programming languages also use escape sequences, and many of them are the
same as in R.


:::{#tip-raw-strings .callout-tip}
A **raw string** is a string where escape sequences are turned off. Raw strings
are especially useful for writing regular expressions (covered in
@sec-regular-expressions).

Raw strings begin with `r"` and an opening delimiter `(`, `[`, or `{`. Raw
strings end with a matching closing delimiter and quote. For example:

```{r}
x = r"(quotes " and backslashes \)"

message(x)
```

Raw strings were added to R in version 4.0 (April 2020), and won't work
correctly in older versions.
:::


#### Character Encodings

Computers store data as numbers. In order to store text on a computer, people
have to agree on a **character encoding**, a system for mapping characters to
numbers. For example, in [ASCII](https://en.wikipedia.org/wiki/ASCII), one of
the most popular encodings in the United States, the character `a` maps to the
number 97.

Many different character encodings exist, and sharing text used to be an
inconvenient process of asking or trying to guess the correct encoding. This
was so inconvenient that in the 1980s, software engineers around the world
united to create the [Unicode](https://home.unicode.org/) standard. Unicode
[includes symbols](http://unicode.org/charts/) for nearly all languages in use
today, as well as emoji and many ancient languages (such as Egyptian
hieroglyphs).

Unicode maps characters to numbers, but unlike a character encoding, it
doesn't dictate how those numbers should be mapped to bytes (sequences of ones
and zeroes). As a result, there are several different character encodings that
support and are synonymous with Unicode. The most popular of these is UTF-8.

In R, you can write Unicode characters with the escape sequence `\u` (or `\U`)
followed by a 4-digit (or 8-digit) number for the character in [base 16][]. For
instance, the number for `a` in Unicode is 97 (the same as in ASCII). In base
16, 97 is `61`. So you can write an `a` as:

```{r}
x = "\u61"
x
```

[base 16]: https://en.wikipedia.org/wiki/Hexadecimal

Unicode escape sequences are usually only used for characters that are not easy
to type. For example, the cat emoji is number `1f408` (in base 16) in Unicode.
So the string `"\U1f408"` is the cat emoji:

```{r}
"\U1f408"
```

:::{.callout-tip}
Whether you can see a Unicode character depends on whether the current font has
a glyph (image representation) for that character. Most fonts only have glyphs
for a few languages.

Make sure to use a font with good Unicode coverage if you love emoji or expect
to work with many different languages. The [Noto Fonts][] project aims to
create a collection of fonts with a common style and complete language
coverage. The [NerdFonts][] project patches fonts commonly used for programming
so that they have better coverage of symbols.

[Noto Fonts]: https://notofonts.github.io/
[NerdFonts]: https://www.nerdfonts.com/
:::

:::{.callout-note}
Most of the time, R will handle character encodings for you automatically.
However, if you ever read or write a text file (including CSV and other
formats) and the text [looks like gibberish][mojibake], it might be an encoding
problem. This is especially true on Windows, the only modern operating system
that does not (yet) use UTF-8 as the default encoding.

[mojibake]: https://en.wikipedia.org/wiki/Mojibake

Encoding problems when reading a file can usually be fixed by passing the
encoding to the function doing the reading. For instance, the code to read a
UTF-8 encoded CSV file on Windows is:

```{r, eval=FALSE}
read.csv("my_data.csv", fileEncoding = "UTF-8")
```

Other reader functions may use a different parameter to set the encoding, so
always check the documentation.

On computers where the native language is not set to English, it can also help
to set R's native language to English with `Sys.setlocale(locale = "English")`.

Encoding problems when writing a file are slightly more complicated to fix. See
[this blog post][ushey-encodings] for thorough explanation.

[ushey-encodings]: https://kevinushey.github.io/blog/2018/02/21/string-encoding-and-r/
:::


### The stringr Package {#sec-the-stringr-package}

Although R has built-in functions for string processing, we recommend using the
[stringr][] package for all of your string processing needs. The package is
part of the Tidyverse, a collection of packages introduced in
@sec-the-tidyverse. Major advantages of stringr over other packages and R's
built-in functions include:

* Correctness: the package builds on [International Components for Unicode
  (ICU)][icu], the Unicode Consortium's own library for handling text encodings
* Discoverability: every function's name begins with `str_` so they're easy to
  discover, remember, and identify in code
* Interface consistency: the first argument is always the string to process,
  the second argument is always the pattern to match (if applicable)
* Vectorization: most of the functions are vectorized in the first and
  second argument

[icu]: https://en.wikipedia.org/wiki/International_Components_for_Unicode

stringr has detailed [documentation][stringr] and also a
[cheatsheet][stringr-cheat-sheet].

[stringr]: https://stringr.tidyverse.org/
[stringr-cheat-sheet]: https://github.com/rstudio/cheatsheets/blob/master/strings.pdf

The first time you use stringr, you'll have to install it with
`install.packages` (the same as any other package). Then you can load the
package with the `library` function:

```{r}
# install.packages("stringr")
library(stringr)
```

The typical syntax of a stringr function is:

```{r, eval = FALSE}
str_name(string, pattern, ...)
```

Where:

* `name` describes what the function does
* `string` is a string to search within or transform
* `pattern` is a pattern to search for, if applicable
* `...` is additional, function-specific arguments

For example, the `str_detect` function *detects* whether a pattern appears
within a string. The function returns `TRUE` if the pattern is found and
`FALSE` if it isn't:

```{r}
str_detect("hello", "el")

str_detect("hello", "ol")
```

Most of the stringr functions are vectorized in the `string` parameter:

```{r}
str_detect(c("hello", "goodbye", "lo"), "lo")
```

As another example, the `str_sub` function extracts a *substring* from a
string, given the substring's position. The first argument is the string, the
second is the position of the substring's first character, and the third is the
position of the substring's last character:

```{r}
str_sub("You speak of destiny as if it was fixed.", 5, 9)
```

The `str_sub` function is especially useful for extracting data from strings
that have a fixed width (although the [readr][] package's `read_fwf` is usually
a better choice if you have a fixed-width file).

[readr]: https://readr.tidyverse.org/

There are a lot of stringr functions. Five that are especially important and
are explained in this reader are:

* `str_detect`, to test whether a string contains a pattern
* `str_sub`, to extract a substring at a given position from a string
* `str_replace`, to replace or remove parts of a string
* `str_split_fixed`, to split a string into parts
* `str_match`, to extract data from a string

You can find a complete list of functions with examples on the stringr
documentation's [reference page][stringr-ref] and the
[cheatsheet][stringr-cheatsheet].

[stringr-ref]: https://stringr.tidyverse.org/reference/
[stringr-cheatsheet]: https://github.com/rstudio/cheatsheets/blob/master/strings.pdf


### Regular Expressions {#sec-regular-expressions}

The stringr functions use a special language called **regular expressions** or
**regex** to describe patterns in strings. Many other programming languages
also have string processing tools that use regular expressions, so fluency with
regular expressions is a valuable, transferrable skill.

You can use a regular expression to describe a complicated pattern in just a
few characters because some characters, called **metacharacters**, have special
meanings. Metacharacters are usually punctation characters. They are *never*
letters or numbers, which always have their literal meaning.

This table lists some of the most useful metacharacters:

Metacharacter | Meaning
------------- | -------
`.`           | any one character (wildcard)
``\``         | escape character (in both R and regex), see @sec-escape-sequences
`^`           | the beginning of string (not a character)
`$`           | the end of string (not a character)
`[ab]`        | one character, either `'a'` or `'b'`
`[^ab]`       | one character, anything except `'a'` or `'b'`
`?`           | the previous character appears 0 or 1 times
`*`           | the previous character appears 0 or more times
`+`           | the previous character appears 1 or more times
`()`          | make a group
`|`           | match left OR right side (not a character)

@sec-regular-expression-examples provides examples of how most of the
metacharacters work.

:::{.callout-note title="See Also"}
Even more examples are presented in the stringr package's [regular expressions
vignette][stringr-regex]. You can find a complete listing of regex
metacharacters in `?regex` or on the stringr [cheatsheet][stringr-cheatsheet].

[stringr-regex]: https://stringr.tidyverse.org/articles/regular-expressions.html
:::

You can disable regular expressions in a stringr function by calling the
`fixed` function on the pattern. For example, to test whether a string contains
a literal dot `.`:

```{r}
x = c("No dot", "Lotsa dots...")
str_detect(x, fixed("."))
```

:::{.callout-tip}
Make it a habit to call `fixed` anytime you use a pattern that doesn't contain
regex metacharacters. Doing so communicates to the reader that you're not using
regex, prevents bugs, and makes your code more efficient.
:::


### Replacing Parts of Strings

Replacing part of a string is a common string processing task. For instance,
quantitative data often contain non-numeric characters such as commas, currency
symbols, and percent signs. These must be removed before converting to numeric
data types. Replacement and removal go hand-in-hand, since removal is
equivalent to replacing part of a string with the **empty string** `""`.

The `str_replace` function replaces the *first* part of a string that matches a
pattern (from left to right), while the related `str_replace_all` function
replaces *every* part of a string that matches a pattern. Most stringr
functions that do pattern matching come in a pair like this: one to process
only the first match and one to process every match.

As an example, suppose you want to remove commas from a vector of numbers (in
strings) so that you can cast them to a numeric data type with `as.numeric`.
You want to remove *all* of the commas, so `str_replace_all` is the function to
use. The first argument is the string and the second is the pattern. The third
argument is the replacement. In this case, the pattern is `","` and the
replacement is the empty string `""`. So the code to remove the commas and cast
the numbers is:

```{r}
x = c("1,000,000", "525,600", "42")
as.numeric(str_replace_all(x, fixed(","), ""))
```

The `str_replace` function wouldn't work as well for this task, since it only
replaces the first match to the pattern:

```{r}
str_replace(x, fixed(","), "")
```

You can also use these functions to replace or remove longer patterns within
words. For instance, suppose you want to change the word `"dog"` to `"cat"`:

```{r}
x = c("dogs are great, dogs are fun", "dogs are fluffy")
str_replace(x, fixed("dog"), "cat")
str_replace_all(x, fixed("dog"), "cat")
```

As a final example, you can use the replacement functions and a regex pattern
to replace repeated spaces with a single space. This is a good standardization
step if you're working with text. The key is to use the regex quantifier `+`,
which means a character "repeats one or more times" in the pattern, and to use
a single space `" "` as the replacement:

```{r}
x = "This    sentence  has  extra      space."
str_replace_all(x, " +", " ")
```

The `str_squish` function does this as well as removing all whitespace from the
beginning and end of a string.


### Splitting Strings {#sec-splitting-strings}

Distinct data in a text are generally separated by a character like a space or
a comma, to make them easy for people to read. Often these separators also make
the data easy for R to parse. The idea is to split the string into a separate
value at each separator.

The `str_split` function splits a string at each match to a pattern. The
matching characters---that is, the separators---are discarded.

For example, suppose you want to split several numbers separated by commas and
spaces:

```{r}
x = "21, 32.3, 5, 64"
result = str_split(x, ", ")
result
```

The `str_split` function always returns a list with one element for each input
string. Here the list only has one element because `x` only has one element.
You can get the first element with:

```{r}
result[[1]]
```

You then convert the values with `as.numeric`.

To see why the `str_split` function always returns a list, consider what
happens if you try to split two different strings at once:

```{r}
x = c(x, "10, 15, 1.3")
result = str_split(x, ", ")
result
```

Each string has a different number of parts, so the vectors in the result have
different lengths. So a list is the only way to store them.

You can also use the `str_split` function to split a sentence into words. Use
spaces for the split:

```{r}
sentences = c(
  "The wonderful, wonderful cat!",
  "Who is this duke of zill anyway?"
)

words = str_split(sentences, " ")
words
```


:::{.callout-tip}
When you know exactly how many parts you expect a string to have, use the
`str_split_fixed` function instead of `str_split`. It requires a third argument
for the maximum number of splits. Because the number of splits is fixed, the
function returns the results in a matrix rather than a list. For example:

```{r}
x = c("1, 2, 3", "10, 20, 30")
str_split_fixed(x, ", ", 3)
```

The `str_split_fixed` function is often more convenient than `str_split`
because the `n`th piece of each input string is just the `n`th column of the
result.

For example, suppose you want to get the area codes from some phone numbers:

```{r}
phones = c("717-555-3421", "629-555-8902", "903-555-6781")
result = str_split_fixed(phones, "-", 3)

result[, 1]
```
:::


### Extracting Matches {#sec-extracting-matches}

Occasionally, you might need to extract parts of a string in a more complicated
way than string splitting allows. One solution is to write a regular expression
that will match all of the data you want to capture, with parentheses `( )`,
the regex metacharacter for a group, around each distinct value. Then you can
use the `str_match` function to extract the groups. @sec-groups presents some
examples of regex groups.

For example, suppose you want to split an email address into three parts: the
user name, the domain name, and the [top-level domain][tld]. To create a
regular expression that matches email addresses, you can use the `@` and `.` in
the address as anchors. The surrounding characters are generally alphanumeric,
which you can represent with the "word" metacharacter `\w`:

```
\w+@\w+[.]\w+
```

Next, put parentheses `( )` around each part that you want to extract:

```
(\w+)@(\w+)[.](\w+)
```

Finally, use this pattern in `str_match`, adding extra backslashes so that
everything is escaped correctly:

```{r}
x = "datalab@ucdavis.edu"
regex = "(\\w+)@(\\w+)[.](\\w+)"
str_match(x, regex)
```

The function extracts the overall match to the pattern, as well as the match to
each group.

The pattern in this example doesn't work for all possible email addresses,
since user names can contain dots and other characters that are not
alphanumeric. You could generalize the pattern if necessary. The point is that
the `str_match` function and groups provide an extremely flexible way to
extract data from strings.


## Case Study: U.S. Cold Storage

The U.S. Department of Agriculture (USDA) publishes a variety of datasets
online, particularly through its National Agricultural Statistics Service
(NASS). Unfortunately, most of are published in PDF or semi-structured text
format, which makes reading the data into R or other statistical software a
challenge.

The USDA NASS posts [monthly reports][usda-warehouse] about stocks of
agricultural products in refrigerated warehouses. In this case study, you'll
use string processing functions to extract a table of data from the December
2022 report.

[usda-warehouse]: https://usda.library.cornell.edu/concern/publications/pg15bd892

:::{.callout-important}
[Click here][usda-warehouse-dec22] to download the December 2022 Cold Storage
report.

If you haven't already, we recommend you create a directory for this workshop.
In your workshop directory, create a `data/` subdirectory. Download and save
the dataset in the `data/` subdirectory.

[usda-warehouse-dec22]: https://ucdavis.box.com/s/7jvab4ys2rfaisozq9078umx8g8vs953
:::

The goal is to extract the first table, about "Nuts, Dairy Products, Frozen
Eggs, and Frozen Poultry," from the report.

The report is a semi-structured mix of natural language text and fixed-width
tables. As a consequence, most functions for reading tabular data will not work
well on the entire report. You could try to use a function for reading
fixed-width data, such as `read.fwf` or the [readr][] package's `read_fwf` on
only the lines containing a table. Another approach, which is shown here, is to
use string processing functions to find and extract the table.

The `readLines` function reads a text file into a character vector with one
element for each line. This makes the function useful for reading unstructured
or semi-structured text. Use the function to read the report:

```{r}
report = readLines("data/2022-12_cold-storage.txt")
head(report)
```

In the report, tables always begin and end with lines that contain only dashes
`-`. By locating these all-dash lines, you can locate the tables. Like
`str_detect`, the `str_which` function tests whether strings in a vector match
a pattern. The only difference is that `str_which` returns the indexes of the
strings that matched (as if you had called `which`) rather than a logical
vector. Use `str_which` to find the all-dash lines:

```{r}
# The regex means:
#   ^  begining of string
#   -+ one or more dashes
#   $  end of string

dashes = str_which(report, "^-+$")
head(report[dashes], 2)
```

Each table contains three dash lines---one separates the header and body. The
header and body of the first table are:

```{r}
report[dashes[1]:dashes[2]]

bod = report[dashes[2]:dashes[3]]
head(bod)
```

The columns have fixed widths, so extracting the columns is relatively easy
with `str_sub` if you can get the offsets. In the last line of the header, the
columns are separated by colons `:`. Thus you can use the `str_locate_all`
function, which returns the locations of a pattern in a string, to get the
offsets:

```{r}
# The regex means:
#   [^:]+  one or more characters, excluding colons
#   (:|$)  a colon or the end of the line

cols = str_locate_all(report[dashes[2] - 1], "[^:]+(:|$)")
# Like str_split, str_locate_all returns a list
cols = cols[[1]]
cols
```

You can use these offsets with `str_sub` to break a line in the body of the
table into columns:

```{r}
str_sub(bod[6], cols)
```

In order to process every line in the body of the table in one vectorized call,
we have to use `str_sub_all` rather than `str_sub`. The result is a list with
one element for each line (that is, a list of rows), but we want a list of
columns, so we also need to transpose the list with the purrr package's
`list_transpose` function:

```{r}
library("purrr")
tab = str_sub_all(bod, cols)
tab = list_transpose(tab)
```

The columns still contain undesirable punctuation and whitespace, but we can
remove these with `str_replace_all` and `str_squish`. We can use purrr's `map`
function to call these on each column:

```{r}
# The regex means:
#   ,     a comma
#   |     OR
#   [.]*  zero or more literal dots
#   :     a colon
#   $     the end of the line

tab = map(tab, \(col) {
  col = str_replace_all(col, ",|[.]*:$", "")
  str_squish(col)
})
tab[[1]]
```

To turn the list of columns into an actual data frame, we can use the dplyr
package's `bind_cols` function:

```{r}
#| message: false
library("dplyr")
tab = bind_cols(tab)
head(tab, 10)
```

The first few rows and the last row can be removed, since they don't contain
data. Then we can convert the individual columns to appropriate data types:

```{r}
tab = tab[-c(1:3, nrow(tab)), ]
tab[2:7] = map(tab[2:7], as.numeric)
head(tab, 10)
```

The data frame is now sufficiently clean that we could use it for a simple
analysis. Of course, there are many things we could do to improve the extracted
data frame, such as identifying categories and subcategories in the first
column, removing rows that are completely empty, and adding column names. These
entail more string processing and data frame manipulation---if you want to
practice your R skills, try doing them on your own.


## Regular Expression Examples {#sec-regular-expression-examples}

:::{.callout-important}
This section is intended as a reference and is not taught in the live session.
:::

This section provides examples of several different regular expression
metacharacters and other features. Most of the examples use the `str_view`
function, which is especially helpful for testing regular expressions. The
function displays an HTML-rendered version of the string with the first match
highlighted.

The [RegExr][] website is also helpful for testing regular expressions; it
provides an interactive interface where you can write regular expressions and
see where they match a string.

[RegExr]: https://regexr.com/

```{r}
#| echo: false

# Ensure str_view output is rendered as HTML.
str_view = function(...) {
  stringr::str_view(..., html = TRUE)
}
```


### The Wildcard

The regex **wildcard** character is `.` and matches any single character. For
example:

```{r}
x = "dog"
str_view(x, "d.g")
```

By default, regex searches from left to right:

```{r}
str_view(x, ".")
```


### Escape Sequences {#sec-escape-sequences}

Like R, regular expressions can contain escape sequences that begin with a
backslash. These are computed separately and after R escape sequences. The main
use for escape sequences in regex is to turn a metacharacter into a literal
character.

For example, suppose you want to match a literal dot `.`. The regex for a
literal dot is `\.`. Since backslashes in R strings have to be escaped, the R
string for this regex is `"\\.`. For example:

```{r}
str_view("this.string", "\\.")
```

The double backslash can be confusing, and it gets worse if you want to match a
literal backslash. You have to escape the backslash in the regex (because
backslash is the regex escape character) and then also have to escape the
backslashes in R (because backslash is also the R escape character). So to
match a single literal backslash in R, the code is:

```{r}
str_view("this\\that", "\\\\")
```

Raw strings (see @tip-raw-strings) make regular expressions easier to read,
because they make backslashes literal (but they still mark the beginning of an
escape sequence in regex). You can use a raw string to write the above as:

```{r}
str_view(r"(this\that)", r"(\\)")
```

### Anchors

By default, a regex will match anywhere in the string. If you want to force a
match at specific place, use an **anchor**.

The beginning of string anchor is `^`. It marks the beginning of the string,
but doesn't count as a character in the pattern.

For example, suppose you want to match an `a` at the beginning of the string:

```{r}
x = c("abc", "cab")

str_view(x, "a")

str_view(x, "^a")
```

It doesn't make sense to put characters before `^`, since no characters can
come before the beginning of the string.

Likewise, the end of string anchor is `$`. It marks the end of the string, but
doesn't count as a character in the pattern.


### Character Classes

In regex, square brackets `[ ]` denote a **character class**. A character class
matches exactly one character, but that character can be any of the characters
inside of the square brackets. The square brackets themselves don't count as
characters in the pattern.

For example, suppose you want to match `c` followed by either `a` or `t`:

```{r}
x = c("ca", "ct", "cat", "cta")

str_view(x, "c[ta]")
```

You can use a dash `-` in a character class to create a **range**. For example,
to match letters `p` through `z`:

```{r}
str_view(x, "c[p-z]")
```

Ranges also work with numbers and capital letters. To match a literal dash,
place the dash at the end of the character class (instead of between two other
characters), as in `[abc-]`.

Most metacharacters are literal when inside a character class. For example,
`[.]` matches a literal dot.

A hat `^` at the beginning of the character class negates the class. So for
example, `[^abc]` matches any one character _except_ for `a`, `b`, or `c`:

```{r}
str_view("abcdef", "[^abc]")
```


### Quantifiers

**Quantifiers** are metacharacters that affect how many times the preceding
character must appear in a match. The quantifier itself doesn't count as a
character in the match.

For example, the question mark `?` quantifier means the preceding character can
appear 0 or 1 times. In other words, `?` makes the preceding character
optional. For example:

```{r}
x = c("abc", "ab", "ac", "abbc")

str_view(x, "ab?c")
```

The star `*` quantifier means the preceding character can appear 0 or more
times. In other words, `*` means the preceding character can appear any number
of times or not at all. For instance:

```{r}
str_view(x, "ab*c")
```

The plus `+` quantifier means the preceding character must appear 1 or more
times.

Quantifiers are **greedy**, meaning they always match as many characters as
possible. In this example, notice that the pattern matches the entire string,
even though it could also match just `abba`:

```{r}
str_view("abbabbba", ".+a")
```

You can add a question mark `?` after another quantifier to make it non-greedy:

```{r}
str_view("abbabbba", ".+?a")
```


### Groups {#sec-groups}

In regex, parentheses `( )` denote a **group**. The parentheses themselves
don't count as characters in the pattern. Groups are useful for repeating or
extracting specific parts of a pattern (see @sec-extracting-matches).

Quantifiers can act on groups in addition to individual characters. For
example, suppose you want to make the entire substring `", dogs,"` optional in
a pattern, so that both of the test strings in this example match:

```{r}
x = c("cats, dogs, and frogs", "cats and frogs")

str_view(x, "cats(, dogs,)? and frogs")
```
