# Next-level Data Visualization

:::{.callout-note title="Learning Goals" collapse="false"}
After this lesson, you should be able to:

* Explain what the grammar of graphics is
* Explain ggplot2's computational model
* Choose appropriate visualizations based on data type(s)
* List the steps to making a well-designed visualization
* Assess what aspects of a visualization need improvement
* Create well-designed data visualizations
:::

This chapter is about creating and customizing data visualizations in R with
the [ggplot2][] package. The first section is an overview of data visalization
systems for R. The second and third sections provide a detailed example of how
to use ggplot2 to recreate a plot. These serve as a review of (or introduction
to) ggplot2's interfaces and computational model; they explain how to reason
about and modify visualizations created with the package. The fourth section
outlines the (non-programming) graphic design skills necessary to make
high-quality plots. Finally, the remaining sections put all of these ideas into
practice through case studies with real data.

[ggplot2]: https://ggplot2.tidyverse.org/


## Data Visualization in R

There are many ways to create visualizations in R, because visualization is a
highly effective way to summarize and gain insights about data. Tools for
creating static visualizations include:

* R's built-in base and graphics packages provide many data visualization
  functions. For instance, the `plot` function creates scatter plots and line
  plots, while the `barplot`, `boxplot`, `histogram`, and `smoothScatter`
  functions are named after the plots they create. With these functions, adding
  more details to a plot is relatively straightforward (for example, use the
  `lines` function to add lines). There's also no need to install any packages
  to use them. A disadvantage is that they all have slightly different
  parameters, which can make them difficult to learn and remember.
* The [tinyplot][] package extends R's built-in data visualization functions to
  provide better support for grouped data, custom color palettes, and more,
  without changing the feel. By design, the package has no other dependencies.
* The [ggplot2][] package uses the grammar of graphics, a convenient way to
  describe visualizations in terms of layers. The package provides a consistent
  interface, detailed [documentation][ggplot2], and a
  [cheatsheet][ggplot2-cheat], all of which are helpful for learning. As R's
  most-downloaded package, ggplot2 is so popular that people have created
  equivalents for the Python and Julia programming languages. The package is
  also part of the [Tidyverse][tidy], a popular collection of R packages
  designed to work well together.
* The [lattice][] package is another package that aims to provide rich support
  for multivariate and grouped data. While lattice creates nice-looking plots
  by default, customizing them usually requires writing a lot of boilerplate
  code.

[tinyplot]: https://grantmcdermott.com/tinyplot/
[ggplot2-cheat]: https://github.com/rstudio/cheatsheets/blob/master/data-visualization.pdf
[tidy]: https://www.tidyverse.org/
[lattice]: https://cran.r-project.org/web/packages/lattice/

Many packages extend these systems further. There are also many packages for
creating dynamic visualizations, which react to user input or change over time,
that we do not cover here.

:::{.callout-warning}
R's built-in data visualization functions, ggplot2, and lattice are not
interoperable! Consequently, it's best to choose one system to use exclusively.

We recommend ggplot2 and will focus on it for the rest of this chapter.
:::

:::{.callout-important}
To follow along with this chapter, you'll need recent versions of the tidyverse
(which includes ggplot2) and palmerpenguins packages. You can install or update
these by running this code in an R prompt:

```{r install-packages, eval=FALSE}
install.packages("tidyverse")
install.packages("palmerpenguins")
```
:::


### The Grammar of Graphics

The "gg" in ggplot2 stands for **grammar of graphics**. The idea of a grammar
of graphics is that visualizations can be built up in layers. In ggplot2, you
can add layers to a plot with plus (`+`). There are three layers that every
plot must have are:

Layer      | Functions | Description
---------- | --------- | -----------
data       | `ggplot2` | Sets the data used to make the plot.
geometry   | `geom_*`  | Sets the geometry (points, lines, etc.) of the plot.
aesthetics | `aes`     | Sets the relationship between data and geometries.

Each layer corresponds to one or more functions. The geometry layer consists of
multiple functions; the asterisk (`*`) in the table is a wildcard.

There are also several optional layers. Here are a few:

Layer       | Functions      | Description
----------- | -------------- | -----------
scale       | `scale_*`      | Sets palettes (color, shape, etc.) and labels
theme       | `theme_*`      | Sets appearance of non-data elements
facets      | `facet_*`      | Makes side-by-side plots
annotation  | `annotation_*` | Adds manually-positioned annotations
guides      | `guide_*`      | Sets axes and legends
coordinate  | `coord_*`      | Sets coordinate systems (Cartesian, logarithmic, polar)

The ggplot2 [documentation][ggplot] provides a complete list of all layers and
their corresponding functions.

:::{.callout-tip}
The grammar of graphics breaks the elements of statistical graphics into parts
in an analogy to human language grammars. Knowing how to put together subject
nouns, object nouns, verbs, and adjectives allows you to construct sentences
that express meaning. Similarly, the grammar of graphics is a collection of
layers and the rules for putting them together to graphically express the
meaning of your data.
:::


## Example: Palmer Penguins, Part 1 {#sec-example-palmer-penguins-part-1}

![
  Artwork by [Allison Horst][horst].
](/images/lter_penguins.png){#fig-penguins}

[horst]: https://allisonhorst.com/

As a refresher about the fundamentals of ggplot2, let's recreate the scatter
plot in @fig-flipper-bill, which shows flipper length versus bill length for
hundreds of individual penguins from three different species: [Adélie][],
[Chinstrap][], and [Gentoo][].

[Adélie]: https://en.wikipedia.org/wiki/Ad%C3%A9lie_penguin
[Chinstrap]: https://en.wikipedia.org/wiki/Chinstrap_penguin
[Gentoo]: https://en.wikipedia.org/wiki/Gentoo_penguin

![
  A scatter plot of data from the [Palmer Penguins dataset][penguins], by
  Allison Horst.
](/images/flipper-bill.png){#fig-flipper-bill}

[penguins]: https://allisonhorst.github.io/palmerpenguins/

The scatter plot uses the [Palmer Penguins dataset][penguins], which was
collected by [Dr. Kristen Gorman][gorman] at [Palmer Station,
Antarctica][palmer]. The plot was created by [Alison Horst][horst], who also
packaged the data for public use. The package is called palmerpenguins; install
it now if you didn't earlier.

[gorman]: https://www.uaf.edu/cfos/people/faculty/detail/kristen-gorman.php
[palmer]: https://pallter.marine.rutgers.edu/

:::{.callout-tip}
Recreating plots from scratch is a great way to practice and learn new data
visualization skills.
:::

Before you start working on any plot, it's a good idea to take a look at the
data. Use the `library` function to load the palmerpenguins package, which
automatically loads the dataset in a variable called `penguins`. Then use the
`head` function to print just the first few rows:

```{r head-penguins}
library("palmerpenguins")

head(penguins)
```

Let's also check the structural summary with the `str` function:

```{r str-penguins}
str(penguins)
```

The structure of a dataset can affect how hard it is to visualize. The
functions in ggplot2 are designed to work with tidy data, which is tabular data
where:

1. Each observation has its own row.
2. Each feature has its own column.
3. Each value has its own cell.

The `penguins` data appears to satisfy all three, which isn't surprising since
this dataset is packaged for others to use. If there was a structural problem,
we would need to reshape the data (see @sec-tidy-datasets).

The types (or classes) of the columns in a dataset can also affect
visualization. Generally, numerical data should be `integer` or `numeric`,
categorical data should be `factor`, Boolean data should be `logical`, and text
should be `character`. Take extra care to convert dates, times, and other
temporal data to appropriate types, as explained in @sec-parsing-dates-times.
Inappropriate data types can cause visualization functions to fail with an
error, or even worse, produce a plot that misrepresents the data.

The data types in the `penguins` data look okay, so we're ready to start
plotting!


### Layer 1: Data

The first layer every ggplot2 plot requires is a **data layer**, which
specifies the data that will be used to make the plot.

You can create a data layer (and the beginnings of a plot) with the `ggplot`
function.

Let's load the package, then use the `ggplot` function to specify that we're
plotting the `penguins` data:

```{r layer1-data}
library("ggplot2")

ggplot(penguins)
```

The result is a blank plot. We've told ggplot2 what data to use, but still
haven't said how to represent it!


### Layer 2: Geometry {#sec-layer-2-geometry}

The second layer every ggplot2 plot requires is a **geometry layer**. The
geometry layer specifies which geometries---points, lines, and other
marks---will represent the data on the plot.

You can create a geometry layer with any of the ggplot2 functions whose names
begin with `geom_`. Each corresponds to a different kind of geometry and plot.
Add the layer to the plot with a plus sign (`+`).

We want to make a scatter plot, so we'll use the `geom_point` function for the
geometry layer. So the code becomes:

```{r layer2-geometry, error=TRUE}
ggplot(penguins) +
  geom_point()
```

When you run this code, you'll see an error message. The plot is still missing
one layer. Fortunately, ggplot2 tells us exactly what we need to do in the
error message.


### Layer 3: Aesthetics

The third layer every ggplot2 plot requires is an **aesthetics layer**. The
aesthetics layer specifies how to represent the data in the data layer with the
geometries in the geometry layer. In other words, it describes the relationship
between the two.

The `aes` function creates an aesthetics layer. You can add it to the plot like
any other layer.

As we saw in the error message in @sec-layer-2-geometry, the `geom_point`
function requires us to set two aesthetics, `x` and `y`. We'll set them to
`flipper_length_mm` and `bill_length_mm`, since these are on the x- and y-axis,
respectively, of @fig-flipper-bill:

```{r layer3-geometry}
ggplot(penguins) +
  aes(x = flipper_length_mm, y = bill_length_mm) +
  geom_point()
```

Now we've got a plot!

Notice that ggplot2 displays a warning about `missing values or values outside
the scale range`. This warning is a reminder that some of the observations in
the dataset are not displayed on the plot, either because they contain missing
values (`NA`) or because they're outside the limits of the plot's axes. By
default, ggplot2 will adjust the axes to include all observations. So in this
case, the warning means that the `penguins` dataset contains 2 missing values.

In @fig-flipper-bill, the color and shape of each point is determined by the
species. In the `penguins` dataset, the `species` column records the species of
each bird. We can connect the color and shape of the points to the species by
setting the optional `color` and `shape` aesthetics:

```{r penguin-plot-color, message=FALSE, warning=FALSE}
ggplot(penguins) +
  aes(
    x = flipper_length_mm,
    y = bill_length_mm,
    color = species,
    shape = species
  ) +
  geom_point()
```

This version of the copycat plot has roughly the same form as
@fig-flipper-bill, although there's still a long way to go to make it exactly
the same. We'll address the remaining differences in
@sec-example-palmer-penguins-part-2.

<!--
We can begin to better understand the grammar of graphics as we consider this
example. Recognize that we our data visualization conveys information via
several visual channels that express data as visual marks. The geometry
determines how those marks are drawn on the page, which can be set separately
from the mapping. Let's see a couple examples of that:

```{r alternate-penguin-geoms, message=FALSE, warning=FALSE}
# placing plots via gridExtra
library(gridExtra)

# plot the Palmer penguin data with a line geometry
peng_line = ggplot(penguins) +
  aes(x = flipper_length_mm,
      y = bill_length_mm,
      color = species,
      shape = species) +
  geom_line()

# plot the Palmer penguin data with a hex heatmap geometry
peng_hex = ggplot(penguins) +
  aes(x = flipper_length_mm,
      y = bill_length_mm,
      color = species,
      shape = species) +
  geom_hex()

# place the plts side-by-side
grid.arrange(peng_line, peng_hex)
```

You can see how changing the geometry but not the mapping will plot the same
data with a different method. Separating the mapping of features to channels
from the drawing of marks is at the core of the grammar of graphics.

This separation of functions gives `ggplot2` its power by allowing us to
compose a small number of functions to express data in unlimited ways (kind of
like poetry). Recognizing the grammar of graphics allows us to reason in a
consistent way about different kinds of plots, and make intelligent assumptions
about mappings and geometries.
-->


## Example: Palmer Penguins, Part 2 {#sec-example-palmer-penguins-part-2}

There are several details we still need to change in the copycat plot from
@sec-example-palmer-penguins-part-1 to make it match @fig-flipper-bill. In
particular:

* The clouds of points are missing the regression lines, which makes it harder
  to see the trends.
* The colors of the points are different. In this case, they're red, green, and
  blue, which are not colorblind-safe.
* The background color and gridline color are different.
* The legend is off to the side.
* There are no titles, and the labels are column names rather than
  descriptions.

We'll fix these in the subsequent sections. Each section focuses on a single
detail and associated ggplot2 layer.


### Adding Regression Lines (Geometry)

The visualization in  @fig-flipper-bill shows a linear regression line for each
group of penguins. Let's add those to our copycat plot. In ggplot2, regression
lines are just another kind of geometry. Lines smooth out noisy data, so the
geometry is called `geom_smooth`.

By default, `geom_smooth` fits and displays a [LOESS][] or [GAM][] line,
depending on the size of the dataset. It also displays the standard error for
the fitted line.

[LOESS]: https://en.wikipedia.org/wiki/Local_regression
[GAM]: https://en.wikipedia.org/wiki/Generalized_additive_model

You can add more than one geometry to a plot, so we'll add `geom_smooth` to the
plot. In order to match @fig-flipper-bill, we'll set `method = lm` to make it
fit a linear regression and `se = FALSE` to hide the standard error:

```{r pengin-plot-lm}
ggplot(penguins) +
  aes(
    x = flipper_length_mm,
    y = bill_length_mm,
    color = species,
    shape = species
  ) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE)
```

Notice ggplot2 prints a new warning, `non-finite outside the scale range`,
about `geom_smooth`. Like the other warning, this is a reminder that there are
missing values in the dataset.


### Changing the Foreground Colors (Scale)

The colors in our copycat plot are red, green, and blue. People with color
blindness may have trouble distinguishing between these colors, so we can
improve the accessibility of the plot by using a different color palette.
@fig-flipper-bill uses orange, purple, and cyan, which are easier to
distinguish. Let's use those colors.

In ggplot2, the **scale layer** controls palettes (for colors, shapes, line
types, and more), as well as text labels. Most of the functions for the scales
layer have names that begin with `scale_`, although there are a few exceptions.
The second part of the name is usually a related aesthetic. For instance, we
can use `scale_color_manual` to manually set the palette for the `color`
aesthetic.

Allison Horst created @fig-flipper-bill with R and ggplot2. The colors she used
are built-in, named colors: `darkorange`, `purple`, and `cyan4`. We'll use
these to set the `values` parameter (which corresponds to the palette) in
`scale_color_manual`:

```{r penguin-plot-palette, message=FALSE, warning=FALSE}
ggplot(penguins) +
  aes(
    x = flipper_length_mm,
    y = bill_length_mm,
    color = species,
    shape = species
  ) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE) +
  scale_color_manual(values = c("darkorange", "purple", "cyan4"))
```

:::{.callout-tip}
You can use Cynthia Brewer's [Color Brewer website][color-brewer] to select
colorblind-friendly color schemes. Color Brewer is integrated directly into
ggplot2. For instance, the `scale_color_brewer` function can pull a named color
scheme from Color Brewer directly into your plot as the color scale.

[color-brewer]: https://colorbrewer2.org/
:::


### Changing the Background Color (Theme)

In @fig-flipper-bill, the background is white and the grid lines are gray, but
in our copycat plot, the background is gray and the grid lines are white. The
color of the background and grid lines, as well as the other visual details of
a plot that are not related to data, are controlled by ggplot2's **theme
layer**.

We can use the `theme_minimal` function to switch to the minimal theme, which
has a white background and gray grid lines:

```{r penguin-plot-theme, message=FALSE, warning=FALSE}
ggplot(penguins) +
  aes(
    x = flipper_length_mm,
    y = bill_length_mm,
    color = species,
    shape = species
  ) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE) +
  scale_color_manual(values = c("darkorange", "purple", "cyan4")) +
  theme_minimal()
```

Several themes are included with ggplot2, and the community has created even
more as extension packages. See the ggplot2 [documentation][ggplot2] for more
details.


### Moving the Legend (Theme)

The theme of a ggplot2 plot also controls where the legend is positioned. So if
we want to move the legend in the copycat plot to inside the plotting area,
like in @fig-flipper-bill, we have to adjust the theme layer.

We can use the `theme` function to make small adjustments to the theme layer.
In this case, we'll set `legend.position` to move the legend inside the
plotting area. We'll also set `legend.position.inside` to specify where to put
the plot (where `c(0, 0)` is the bottom left and `c(1, 1)` is the top right).
The code for the copycat plot becomes:

```{r penguin-plot-legend, message=FALSE, warning=FALSE}
ggplot(penguins) +
  aes(
    x = flipper_length_mm,
    y = bill_length_mm,
    color = species,
    shape = species
  ) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE) +
  scale_color_manual(values = c("darkorange", "purple", "cyan4")) +
  theme_minimal() +
  theme(legend.position = "inside", legend.position.inside = c(0.85, 0.15))
```

See the ggplot2 [documentation][ggplot2] for the complete list of parameters
you can use to fine-tune the theme.


### Adding Labels (Scale)

Adding labels and titles to a visualization is a critical step. Without labels,
the viewer might not be able to understand what the visualization shows. In
ggplot2, we can set these with the scale layer function `labs` (short for
"labels"). We can set a title, subtitle, and also a label for each aesthetic.

Here's how to set the labels on the copycat plot to match @fig-flipper-bill:

```{r penguin-plot-labels, message=FALSE, warning=FALSE}
ggplot(penguins) +
  aes(
    x = flipper_length_mm,
    y = bill_length_mm,
    color = species,
    shape = species
  ) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE) +
  scale_color_manual(values = c("darkorange", "purple", "cyan4")) +
  theme_minimal() +
  theme(legend.position = "inside", legend.position.inside = c(0.85, 0.15)) +
  labs(
    title = "Flipper and bill length",
    subtitle = "Dimensions for Adelie, Chinstrap and Gentoo Penguins at Palmer Station LTER",
    x = "Flipper length (mm)",
    y = "Bill length (mm)",
    color = "Penguin species",
    shape = "Penguin species"
  )
```


## Visualization Best Practices

Besides programming skills, to create a visualization that conveys a clear
message about a dataset, you need to know graphic design best practices and the
principles of visual perception. DataLab's [Principles of Data
Visualization][datalab-data-viz] covers these important skills.

[datalab-data-viz]: https://ucdavisdatalab.github.io/workshop_data_viz_principles/

DataLab's [Data Visualization Guidelines][checklist] is a concise reference and
reminder of these skills. Bookmark or print out a copy and run through the
checklist whenever you design a visualization. The case studies in the
{numref}`case-studies` show how to fix some of the issues on the checklist.

[checklist]: /docs/graphics_checklist.pdf

:::{.callout-note title="Revisiting the Palmer Penguins Plot"}
Let's apply the Data Visualization Guidelines to the Palmer Penguins plot
(@fig-flipper-bill):

* This plot has two numerical features (bill length and flipper length),
  expressed using a scatter plot. There's also a categorical feature (species),
  which is indicated by the different colors and shapes of the plot.
* The plot expresses the fact that flipper length is positively associated with
  bill length for all three species of penguins, but the sizes and the
  relationships between them are unique to each species.
* There's a title and a legend, the axes are labeled with units, and all of
  the text is in plain language.
* There's a risk that the data may hide the message, so a smoothing line is
  added to each species for clarity.
* The colors are accessible to colorblind people.

This is a good data visualization!
:::


## Case Study: Counting Penguins

First, let's revisit the `penguins` data. There are tree categorical features
in the data: `species`, `island`, and `sex`. Let's use `geom_bar` to count how
many penguins of each species and/or sex were observed on each island. The
x-axis of the plot should be the island, but note that there are multiple
values of `species` and `sex` that have the same position on that x-axis. In
this case, you can use the `position_dodge` or `position_stack` arguments to
tell `ggplot2` how to handle the second grouping channel.

```{r counting-penguins, warning=FALSE, message=FALSE}
# count the penguins on each island
ggplot(penguins) +
  aes(x=island) +
  geom_bar() +
  xlab("Island") +
  ylab("Count") +
  ggtitle("Count of penguins on each island")


#count the penguins of each sex on each island
ggplot(penguins) +
  aes(x=island, fill=sex) +
  geom_bar(position=position_dodge()) +
  scale_fill_grey() +
  theme_bw() +
  xlab("Island") +
  ylab("Count") +
  ggtitle("Count of penguins on each island by sex")
```

Alternatively, you can use facets to separate the data into multiple plots
based on a data feature. Let's see how that works to facet the plots by
species.

One way to show more information more clearly in a plot is to break the plot
into pieces that each show part of the information. In `ggplot2`, this is
called **faceting** the plot. There are two main facet functions, `facet_grid`
(which puts plots in a grid), and `facet_wrap`, which puts plots side-by-side
until it runs out of room, then wraps to a new line. We'll use `facet_wrap`
here, with the first argument being `~species`. This tells `ggplot2` to break
the plot into pieces by plotting the data for each `species` separately.

```{r penguin-facet-species, warning=FALSE, message=FALSE}
#count the penguins of each species on each island
ggplot(penguins) +
  aes(x=island) +
  geom_bar() +
  scale_fill_grey() +
  theme_bw() +
  xlab("Island") +
  ylab("Count") +
  ggtitle("Count of penguins on each island by species") +
  facet_wrap(~species, ncol=3)
```


## Case Study: Error Bars

Here's an example that recently came up in my office hours. You've done an
experiment to see how mice with two different genotypes respond to two
different treatments. Now you want to plot the mean response of each group as a
column, with error bars indicating the standard deviation of the mean. You also
want to show the raw data. I've simulated some data for us to use---download it
[here](https://ucdavis.box.com/s/lxgplwqn23ncbeb5gc6ktpfknr56xws6).

This one is kind of complicated because you have to tell `ggplot2` how to
calculate the height of the columns and of the error bars.

```{r example-genotype, warning=FALSE, message=FALSE}
library("readr")

mice_geno = read_csv("data/genotype-response.csv")

# show the treatment response for different genotypes
ggplot(mice_geno) +
  aes(x=trt,
      y=resp,
      fill=genotype) +
  scale_fill_brewer(palette="Dark2") +
  geom_bar(position=position_dodge(),
           stat='summary',
           fun='mean') +
  geom_errorbar(fun.min=function(x) {mean(x) - sd(x) / sqrt(length(x))},
               fun.max=function(x) {mean(x) + sd(x) / sqrt(length(x))},
               stat='summary',
               position=position_dodge(0.9),
               width=0.2) +
  geom_point(position=
               position_jitterdodge(
                 dodge.width=0.9,
                 jitter.width=0.1)) +
  xlab("Treatment") +
  ylab("Response (mm/g)") +
  ggtitle("Mean growth response of mice by genotype and treatment")
```


## Case Study: Bird Flu Mortality

People mail dead birds to the USDA and USGS, where scientists analyze the birds
to find out why they died. Right now there is a bird flu epidemic, and the USDA
provides public data about the birds in whom the disease has been detected. You
can access the data
[here](https://ucdavis.box.com/s/babz1b5ak0kh7q7cktg7k5gzrt2ve9gc) or see the
official USDA webpage
[here](https://www.aphis.usda.gov/aphis/ourfocus/animalhealth/animal-disease-information/avian/avian-influenza/).
After you download the data, we will load the data and do some visualization.

```{r example-bird-flu, message=FALSE, warning=FALSE}
library("lubridate")

# load data directly from the USDA website
flu <- read_csv("data/hpai-wild-birds-ver2.csv")
flu$date <- mdy(flu$`Date Detected`)

# plot a histogram of when bird flu was detected
ggplot(flu) +
  aes(x = date) +
  geom_histogram() +
  ggtitle("Bird flu detections in wild birds") +
  xlab("Date") +
  ylab("Count")

# plot a histogram of when bird flu was detected
ggplot(flu) +
  aes(x = date, fill = `Sampling Method`) +
  geom_histogram() +
  ggtitle("Bird flu detections in wild birds") +
  xlab("Date") +
  ylab("Count")

# bar chart shows how the bird flu reports compare between west coast states
subset(flu, State %in% c("California", "Oregon", "Washington")) |>
  ggplot() +
  aes(x = State, fill = `Sampling Method`) +
  stat_count() +
  geom_bar() +
  ggtitle("Bird flu detections in wild birds (West coast states)") +
  ylab("Count")
```


Let's compare the bird flu season to the human flu season. Download
hospitalization data for the 2021-2022 and 2022-2023 flu seasons from the CDC
website [here](https://gis.cdc.gov/GRASP/Fluview/FluHospRates.html) or see the
official Centers for Disease Control website
[here](https://gis.cdc.gov/GRASP/Fluview/FluHospRates.html). After you download
the data, we will see how adding a second data series works a little
differently from the first. That's because composing data, aesthetic mapping,
and geometry with addition only works when there is no ambiguity about which
data series is being mapped or drawn.

After downloading the data, there is some work required to adjust the dates and
change the hospitalization rate from cases per 100,000 to cases per 10 million,
which better matches the scale of the bird flu data.

```{r example-flu-processing, message=FALSE, warning=FALSE}
# processing CDC flu data:
cdc <- read_csv("data/FluSurveillance_Custom_Download_Data.csv", skip = 2)
cdc$date <- as_date("1950-01-01")
year(cdc$date) <- cdc$`MMWR-YEAR`
week(cdc$date) <- cdc$`MMWR-WEEK`


# get flu hospitalization counts that include all race, sex, and age categories
cdc_overall <- subset(
  cdc,
  `AGE CATEGORY` == "Overall" &
    `SEX CATEGORY` == "Overall" &
    `RACE CATEGORY` == "Overall"
)

# convert the counts to cases per 10 million
cdc_overall$`WEEKLY RATE` <- as.numeric(cdc_overall$`WEEKLY RATE`) * 100
```


```{r example-flu-plotting, warning=FALSE, message=FALSE}
# remake the plot but add a new geom_line() with its own data
ggplot(flu) +
  aes(x = date, fill = `Sampling Method`) +
  geom_histogram() +
  geom_line(data = cdc_overall,
            mapping = aes(x = date, y = `WEEKLY RATE`),
            inherit.aes = FALSE) +
  ggtitle("Bird flu detections and human flu hospitalizations") +
  xlab("Date") +
  ylab("Count") +
  xlim(as_date("2022-01-01"), as_date("2023-05-01"))
```


## Case Study: Small Business Loans

The US Small Business Administration (SBA) maintains data on the loans it
offers to businesses. Data about loans made since 2020 can be found at the
[Small Business Administration
website](https://data.sba.gov/dataset/7-a-504-foia), or you can download it
from [here](https://ucdavis.box.com/s/ctx6grx10c83lh561ekl9tnmi20561ed). We'll
load that data and then explore some ways to visualize it. Since the difference
between a \$100 loan and a \$1000 loan is more like the difference between
\$100,000 and \$1M than between \$100,000 ad $100,900, we should put the loan
values on a  logarithmic scale. You can do this in `ggplot2` with the
`scale_y_log10` function (when the loan values are on the y axis).


```{r example-SBA, message=FALSE, warning=FALSE}
# load the small business loan data
sba <- read_csv("data/small-business-loans.csv")

# check the SBA data to see the data types, etc.
head(sba)

# boxplot of loan sizes by business type
subset(sba, ProjectState == "CA") |>
  ggplot() +
  aes(x = BusinessType, y = SBAGuaranteedApproval) +
  geom_boxplot() +
  scale_y_log10() +
  ggtitle("Small Business Administraton guaranteed loans in California") +
  ylab("Loan guarantee (dollars)")


# relationship between loan size and interest rate
subset(sba, ProjectState == "CA") |>
  ggplot() +
  aes(x = GrossApproval, y = InitialInterestRate, ) +
  geom_point() +
  facet_wrap(~BusinessType, ncol = 3) +
  scale_x_log10() +
  ggtitle("Interest rate as a function of loan size") +
  xlab("Loan size (dollars)") +
  ylab("Interest rate (%)")
```


Now let's color the points by the loan status. Thankfully, `ggplot2` integrates
directly with [Color Brewer (colorbrewer2.org)](https://colorbrewer2.org) to
get better color palettes. We will use the `Accent` color palette, which is
just one of the many options that can be found on the Color Brewer site.

There are a lot of data points, which tent to largely overlap and hide each
other. We use a smoother (`geom_smooth`) to help call out differences that
would otherwise be lost in the noise of the points.

```{r example-SBA-facet, message=FALSE, warning=FALSE}
# color the dots by the loan status.
subset(sba, ProjectState == "CA" & LoanStatus != "EXEMPT" & LoanStatus != "CHGOFF") |>
  ggplot() +
  aes(x = GrossApproval, y = InitialInterestRate, color = LoanStatus) +
  geom_point() +
  geom_smooth() +
  facet_wrap(~BusinessType, ncol = 3) +
  scale_x_log10() +
  ggtitle("Interest rate as a function of loan size by loan status") +
  xlab("Loan size (dollars)") +
  ylab("Interest rate (%)") +
  labs(color = "Loan status") +
  scale_color_brewer(type = "qual", palette = "Accent")
```
