# Next-level Data Visualization

:::{.callout-note title="Learning Goals" collapse="false"}
After this lesson, you should be able to:

* Explain what the grammar of graphics is
* Explain ggplot2's computational model
* Choose appropriate visualizations based on data type(s)
* List the steps to making a well-designed visualization
* Assess what aspects of a visualization need improvement
* Create well-designed data visualizations
:::

This chapter is about creating and customizing data visualizations in R with
the [ggplot2][] package. The first section is an overview of data visalization
systems for R. The second and third sections provide a detailed example of how
to use ggplot2 to recreate a plot. These serve as a review of (or introduction
to) ggplot2's interfaces and computational model; they explain how to reason
about and modify visualizations created with the package. The fourth section
outlines the (non-programming) graphic design skills necessary to make
high-quality plots. Finally, the remaining sections put all of these ideas into
practice through case studies with real data.

[ggplot2]: https://ggplot2.tidyverse.org/


## Data Visualization in R

There are many ways to create visualizations in R, because visualization is a
highly effective way to summarize and gain insights about data. Tools for
creating static visualizations include:

* R's built-in base and graphics packages provide many data visualization
  functions. For instance, the `plot` function creates scatter plots and line
  plots, while the `barplot`, `boxplot`, `histogram`, and `smoothScatter`
  functions are named after the plots they create. With these functions, adding
  more details to a plot is relatively straightforward (for example, use the
  `lines` function to add lines). There's also no need to install any packages
  to use them. A disadvantage is that they all have slightly different
  parameters, which can make them difficult to learn and remember.
* The [tinyplot][] package extends R's built-in data visualization functions to
  provide better support for grouped data, custom color palettes, and more,
  without changing the feel. By design, the package has no other dependencies.
* The [ggplot2][] package uses the grammar of graphics, a convenient way to
  describe visualizations in terms of layers. The package provides a consistent
  interface, detailed [documentation][ggplot2], and a
  [cheatsheet][ggplot2-cheat], all of which are helpful for learning. As R's
  most-downloaded package, ggplot2 is so popular that people have created
  equivalents for the Python and Julia programming languages. The package is
  also part of the [Tidyverse][tidy], a popular collection of R packages
  designed to work well together.
* The [lattice][] package is another package that aims to provide rich support
  for multivariate and grouped data. While lattice creates nice-looking plots
  by default, customizing them usually requires writing a lot of boilerplate
  code.

[tinyplot]: https://grantmcdermott.com/tinyplot/
[ggplot2-cheat]: https://github.com/rstudio/cheatsheets/blob/master/data-visualization.pdf
[tidy]: https://www.tidyverse.org/
[lattice]: https://cran.r-project.org/web/packages/lattice/

Many packages extend these systems further. There are also many packages for
creating dynamic visualizations, which react to user input or change over time,
that we do not cover here.

:::{.callout-warning}
R's built-in data visualization functions, ggplot2, and lattice are not
interoperable! Consequently, it's best to choose one system to use exclusively.

We recommend ggplot2 and will focus on it for the rest of this chapter.
:::

:::{.callout-important}
To follow along with this chapter, you'll need recent versions of the tidyverse
(which includes ggplot2) and palmerpenguins packages. You can install or update
these by running this code in an R prompt:

```{r install-packages, eval=FALSE}
install.packages("tidyverse")
install.packages("palmerpenguins")
```
:::


### The Grammar of Graphics

The "gg" in ggplot2 stands for **grammar of graphics**. The idea of a grammar
of graphics is that visualizations can be built up in layers. In ggplot2, you
can add layers to a plot with plus (`+`). There are three layers that every
plot must have are:

Layer      | Functions | Description
---------- | --------- | -----------
data       | `ggplot2` | Sets the data used to make the plot.
geometry   | `geom_*`  | Sets the geometry (points, lines, etc.) of the plot.
aesthetics | `aes`     | Sets the relationship between data and geometries.

Each layer corresponds to one or more functions. The geometry layer consists of
multiple functions; the asterisk (`*`) in the table is a wildcard.

There are also several optional layers. Here are a few:

Layer       | Functions      | Description
----------- | -------------- | -----------
scale       | `scale_*`      | Sets palettes (color, shape, etc.) and labels
theme       | `theme_*`      | Sets appearance of non-data elements
facets      | `facet_*`      | Makes side-by-side plots
annotation  | `annotation_*` | Adds manually-positioned annotations
guides      | `guide_*`      | Sets axes and legends
coordinate  | `coord_*`      | Sets coordinate systems (Cartesian, logarithmic, polar)

The ggplot2 [documentation][ggplot] provides a complete list of all layers and
their corresponding functions.

:::{.callout-tip}
The grammar of graphics breaks the elements of statistical graphics into parts
in an analogy to human language grammars. Knowing how to put together subject
nouns, object nouns, verbs, and adjectives allows you to construct sentences
that express meaning. Similarly, the grammar of graphics is a collection of
layers and the rules for putting them together to graphically express the
meaning of your data.
:::


## Example: Palmer Penguins, Part 1 {#sec-example-palmer-penguins-part-1}

![
  Artwork by [Allison Horst][horst].
](/images/lter_penguins.png){#fig-penguins}

[horst]: https://allisonhorst.com/

As a refresher about the fundamentals of ggplot2, let's recreate the scatter
plot in @fig-flipper-bill, which shows flipper length versus bill length for
hundreds of individual penguins from three different species: [Adélie][],
[Chinstrap][], and [Gentoo][].

[Adélie]: https://en.wikipedia.org/wiki/Ad%C3%A9lie_penguin
[Chinstrap]: https://en.wikipedia.org/wiki/Chinstrap_penguin
[Gentoo]: https://en.wikipedia.org/wiki/Gentoo_penguin

![
  A scatter plot of data from the [Palmer Penguins dataset][penguins], by
  Allison Horst.
](/images/flipper-bill.png){#fig-flipper-bill}

[penguins]: https://allisonhorst.github.io/palmerpenguins/

The scatter plot uses the [Palmer Penguins dataset][penguins], which was
collected by [Dr. Kristen Gorman][gorman] at [Palmer Station,
Antarctica][palmer]. The plot was created by [Alison Horst][horst], who also
packaged the data for public use. The package is called palmerpenguins; install
it now if you didn't earlier.

[gorman]: https://www.uaf.edu/cfos/people/faculty/detail/kristen-gorman.php
[palmer]: https://pallter.marine.rutgers.edu/

:::{.callout-tip}
Recreating plots from scratch is a great way to practice and learn new data
visualization skills.
:::

Before you start working on any plot, it's a good idea to take a look at the
data. Use the `library` function to load the palmerpenguins package, which
automatically loads the dataset in a variable called `penguins`. Then use the
`head` function to print just the first few rows:

```{r head-penguins}
library("palmerpenguins")

head(penguins)
```

Let's also check the structural summary with the `str` function:

```{r str-penguins}
str(penguins)
```

The structure of a dataset can affect how hard it is to visualize. The
functions in ggplot2 are designed to work with tidy data, which is tabular data
where:

1. Each observation has its own row.
2. Each feature has its own column.
3. Each value has its own cell.

The `penguins` data appears to satisfy all three, which isn't surprising since
this dataset is packaged for others to use. If there was a structural problem,
we would need to reshape the data (see @sec-tidy-datasets).

The types (or classes) of the columns in a dataset can also affect
visualization. Generally, numerical data should be `integer` or `numeric`,
categorical data should be `factor`, Boolean data should be `logical`, and text
should be `character`. Take extra care to convert dates, times, and other
temporal data to appropriate types, as explained in @sec-parsing-dates-times.
Inappropriate data types can cause visualization functions to fail with an
error, or even worse, produce a plot that misrepresents the data.

The data types in the `penguins` data look okay, so we're ready to start
plotting!


### Layer 1: Data

The first layer every ggplot2 plot requires is a **data layer**, which
specifies the data that will be used to make the plot.

You can create a data layer (and the beginnings of a plot) with the `ggplot`
function.

Let's load the package, then use the `ggplot` function to specify that we're
plotting the `penguins` data:

```{r layer1-data}
library("ggplot2")

ggplot(penguins)
```

The result is a blank plot. We've told ggplot2 what data to use, but still
haven't said how to represent it!


### Layer 2: Geometry {#sec-layer-2-geometry}

The second layer every ggplot2 plot requires is a **geometry layer**. The
geometry layer specifies which geometries---points, lines, and other
marks---will represent the data on the plot.

You can create a geometry layer with any of the ggplot2 functions whose names
begin with `geom_`. Each corresponds to a different kind of geometry and plot.
Add the layer to the plot with a plus sign (`+`).

We want to make a scatter plot, so we'll use the `geom_point` function for the
geometry layer. So the code becomes:

```{r layer2-geometry, error=TRUE}
ggplot(penguins) +
  geom_point()
```

When you run this code, you'll see an error message. The plot is still missing
one layer. Fortunately, ggplot2 tells us exactly what we need to do in the
error message.


### Layer 3: Aesthetics

The third layer every ggplot2 plot requires is an **aesthetics layer**. The
aesthetics layer specifies how to represent the data in the data layer with the
geometries in the geometry layer. In other words, it describes the relationship
between the two.

The `aes` function creates an aesthetics layer. You can add it to the plot like
any other layer.

As we saw in the error message in @sec-layer-2-geometry, the `geom_point`
function requires us to set two aesthetics, `x` and `y`. We'll set them to
`flipper_length_mm` and `bill_length_mm`, since these are on the x- and y-axis,
respectively, of @fig-flipper-bill:

```{r layer3-geometry}
ggplot(penguins) +
  aes(x = flipper_length_mm, y = bill_length_mm) +
  geom_point()
```

Now we've got a plot!

Notice that ggplot2 displays a warning about `missing values or values outside
the scale range`. This warning is a reminder that some of the observations in
the dataset are not displayed on the plot, either because they contain missing
values (`NA`) or because they're outside the limits of the plot's axes. By
default, ggplot2 will adjust the axes to include all observations. So in this
case, the warning means that the `penguins` dataset contains 2 missing values.

In @fig-flipper-bill, the color and shape of each point is determined by the
species. In the `penguins` dataset, the `species` column records the species of
each bird. We can connect the color and shape of the points to the species by
setting the optional `color` and `shape` aesthetics:

```{r penguin-plot-color, message=FALSE, warning=FALSE}
ggplot(penguins) +
  aes(
    x = flipper_length_mm,
    y = bill_length_mm,
    color = species,
    shape = species
  ) +
  geom_point()
```

This version of the copycat plot has roughly the same form as
@fig-flipper-bill, although there's still a long way to go to make it exactly
the same. We'll address the remaining differences in
@sec-example-palmer-penguins-part-2.

<!--
We can begin to better understand the grammar of graphics as we consider this
example. Recognize that we our data visualization conveys information via
several visual channels that express data as visual marks. The geometry
determines how those marks are drawn on the page, which can be set separately
from the mapping. Let's see a couple examples of that:

```{r alternate-penguin-geoms, message=FALSE, warning=FALSE}
# placing plots via gridExtra
library(gridExtra)

# plot the Palmer penguin data with a line geometry
peng_line = ggplot(penguins) +
  aes(x = flipper_length_mm,
      y = bill_length_mm,
      color = species,
      shape = species) +
  geom_line()

# plot the Palmer penguin data with a hex heatmap geometry
peng_hex = ggplot(penguins) +
  aes(x = flipper_length_mm,
      y = bill_length_mm,
      color = species,
      shape = species) +
  geom_hex()

# place the plts side-by-side
grid.arrange(peng_line, peng_hex)
```

You can see how changing the geometry but not the mapping will plot the same
data with a different method. Separating the mapping of features to channels
from the drawing of marks is at the core of the grammar of graphics.

This separation of functions gives `ggplot2` its power by allowing us to
compose a small number of functions to express data in unlimited ways (kind of
like poetry). Recognizing the grammar of graphics allows us to reason in a
consistent way about different kinds of plots, and make intelligent assumptions
about mappings and geometries.
-->


## Example: Palmer Penguins, Part 2 {#sec-example-palmer-penguins-part-2}

There are several details we still need to change in the copycat plot from
@sec-example-palmer-penguins-part-1 to make it match @fig-flipper-bill. In
particular:

* The clouds of points are missing the regression lines, which makes it harder
  to see the trends.
* The colors of the points are different. In this case, they're red, green, and
  blue, which are not colorblind-safe.
* The background color and gridline color are different.
* The legend is off to the side.
* There are no titles, and the labels are column names rather than
  descriptions.

We'll fix these in the subsequent sections. Each section focuses on a single
detail and associated ggplot2 layer.


### Adding Regression Lines (Geometry)

The visualization in  @fig-flipper-bill shows a linear regression line for each
group of penguins. Let's add those to our copycat plot. In ggplot2, regression
lines are just another kind of geometry. Lines smooth out noisy data, so the
geometry is called `geom_smooth`.

By default, `geom_smooth` fits and displays a [LOESS][] or [GAM][] line,
depending on the size of the dataset. It also displays the standard error for
the fitted line.

[LOESS]: https://en.wikipedia.org/wiki/Local_regression
[GAM]: https://en.wikipedia.org/wiki/Generalized_additive_model

You can add more than one geometry to a plot, so we'll add `geom_smooth` to the
plot. In order to match @fig-flipper-bill, we'll set `method = lm` to make it
fit a linear regression and `se = FALSE` to hide the standard error:

```{r pengin-plot-lm}
ggplot(penguins) +
  aes(
    x = flipper_length_mm,
    y = bill_length_mm,
    color = species,
    shape = species
  ) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE)
```

Notice ggplot2 prints a new warning, `non-finite outside the scale range`,
about `geom_smooth`. Like the other warning, this is a reminder that there are
missing values in the dataset.


### Changing the Foreground Colors (Scale)

The colors in our copycat plot are red, green, and blue. People with color
blindness may have trouble distinguishing between these colors, so we can
improve the accessibility of the plot by using a different color palette.
@fig-flipper-bill uses orange, purple, and cyan, which are easier to
distinguish. Let's use those colors.

In ggplot2, the **scale layer** controls palettes (for colors, shapes, line
types, and more), as well as text labels. Most of the functions for the scales
layer have names that begin with `scale_`, although there are a few exceptions.
The second part of the name is usually a related aesthetic. For instance, we
can use `scale_color_manual` to manually set the palette for the `color`
aesthetic.

Allison Horst created @fig-flipper-bill with R and ggplot2. The colors she used
are built-in, named colors: `darkorange`, `purple`, and `cyan4`. We'll use
these to set the `values` parameter (which corresponds to the palette) in
`scale_color_manual`:

```{r penguin-plot-palette, message=FALSE, warning=FALSE}
ggplot(penguins) +
  aes(
    x = flipper_length_mm,
    y = bill_length_mm,
    color = species,
    shape = species
  ) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE) +
  scale_color_manual(values = c("darkorange", "purple", "cyan4"))
```

:::{.callout-tip}
You can use Cynthia Brewer's [Color Brewer website][color-brewer] to select
colorblind-friendly color schemes. Color Brewer is integrated directly into
ggplot2. For instance, the `scale_color_brewer` function can pull a named color
scheme from Color Brewer directly into your plot as the color scale.

[color-brewer]: https://colorbrewer2.org/
:::


### Changing the Background Color (Theme)

In @fig-flipper-bill, the background is white and the grid lines are gray, but
in our copycat plot, the background is gray and the grid lines are white. The
color of the background and grid lines, as well as the other visual details of
a plot that are not related to data, are controlled by ggplot2's **theme
layer**.

We can use the `theme_minimal` function to switch to the minimal theme, which
has a white background and gray grid lines:

```{r penguin-plot-theme, message=FALSE, warning=FALSE}
ggplot(penguins) +
  aes(
    x = flipper_length_mm,
    y = bill_length_mm,
    color = species,
    shape = species
  ) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE) +
  scale_color_manual(values = c("darkorange", "purple", "cyan4")) +
  theme_minimal()
```

Several themes are included with ggplot2, and the community has created even
more as extension packages. See the ggplot2 [documentation][ggplot2] for more
details.


### Moving the Legend (Theme)

The theme of a ggplot2 plot also controls where the legend is positioned. So if
we want to move the legend in the copycat plot to inside the plotting area,
like in @fig-flipper-bill, we have to adjust the theme layer.

We can use the `theme` function to make small adjustments to the theme layer.
In this case, we'll set `legend.position` to move the legend inside the
plotting area. We'll also set `legend.position.inside` to specify where to put
the plot (where `c(0, 0)` is the bottom left and `c(1, 1)` is the top right).
The code for the copycat plot becomes:

```{r penguin-plot-legend, message=FALSE, warning=FALSE}
ggplot(penguins) +
  aes(
    x = flipper_length_mm,
    y = bill_length_mm,
    color = species,
    shape = species
  ) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE) +
  scale_color_manual(values = c("darkorange", "purple", "cyan4")) +
  theme_minimal() +
  theme(legend.position = "inside", legend.position.inside = c(0.85, 0.15))
```

See the ggplot2 [documentation][ggplot2] for the complete list of parameters
you can use to fine-tune the theme.


### Adding Labels (Scale)

Adding labels and titles to a visualization is a critical step. Without labels,
the viewer might not be able to understand what the visualization shows. In
ggplot2, we can set these with the scale layer function `labs` (short for
"labels"). We can set a title, subtitle, and also a label for each aesthetic.

Here's how to set the labels on the copycat plot to match @fig-flipper-bill:

```{r penguin-plot-labels, message=FALSE, warning=FALSE}
ggplot(penguins) +
  aes(
    x = flipper_length_mm,
    y = bill_length_mm,
    color = species,
    shape = species
  ) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE) +
  scale_color_manual(values = c("darkorange", "purple", "cyan4")) +
  theme_minimal() +
  theme(legend.position = "inside", legend.position.inside = c(0.85, 0.15)) +
  labs(
    title = "Flipper and bill length",
    subtitle = "Dimensions for Adelie, Chinstrap and Gentoo Penguins at Palmer Station LTER",
    x = "Flipper length (mm)",
    y = "Bill length (mm)",
    color = "Penguin species",
    shape = "Penguin species"
  )
```


## Visualization Best Practices

Besides programming skills, to create a visualization that conveys a clear
message about a dataset, you need to know graphic design best practices and the
principles of visual perception. DataLab's [Principles of Data
Visualization][datalab-data-viz] covers these important skills.

[datalab-data-viz]: https://ucdavisdatalab.github.io/workshop_data_viz_principles/

DataLab's [Data Visualization Guidelines][checklist] is a concise reference and
reminder of these skills. Bookmark or print out a copy and run through the
checklist whenever you design a visualization. The case studies in the
{numref}`case-studies` show how to fix some of the issues on the checklist.

[checklist]: /docs/graphics_checklist.pdf

:::{.callout-note title="Revisiting the Palmer Penguins Plot"}
Let's apply the Data Visualization Guidelines to the Palmer Penguins plot
(@fig-flipper-bill):

* This plot has two numerical features (bill length and flipper length),
  expressed using a scatter plot. There's also a categorical feature (species),
  which is indicated by the different colors and shapes of the plot.
* The plot expresses the fact that flipper length is positively associated with
  bill length for all three species of penguins, but the sizes and the
  relationships between them are unique to each species.
* There's a title and a legend, the axes are labeled with units, and all of
  the text is in plain language.
* There's a risk that the data may hide the message, so a smoothing line is
  added to each species for clarity.
* The colors are accessible to colorblind people.

This is a good data visualization!
:::


## Case Study: Plotting a Categorical Distribution {#sec-plotting-categorical-distribution}

:::{.callout-note}
This case study shows how to:

* Plot the distribution of one or more categorical features
* Create faceted plots
:::

When you start working with a new dataset, the first thing you should do is
explore (and clean) the features. Exploring the features means inspecting the
distribution of each feature, as well as checking for relationships between
features. For this case study, let's use ggplot2 to explore the categorical
features in the Palmer Penguins data set introduced in
@sec-example-palmer-penguins-part-1.

Consider the species of the penguins the data set. Are the three
species---Adélie, Chinstrap, and Gentoo---equally represented? This question is
asking how the categorical `species` feature is distributed. Tables, bar plots,
and dot plots are all good tools for examining the distribution of a single
categorical feature. Let's make a table and a bar plot.

:::{.callout-tip}
Inspecting data through multiple methods is a good way to verify that your code
and interpretations are correct.
:::

You can use the R's `table` function to make a table:

```{r penguins-table}
table(penguins$species)
```

To make the bar plot, you can use ggplot2's `geom_bar` function. Make a bar
plot of the distribution of penguin species:

```{r penguins-bar-species}
ggplot(penguins) +
  aes(x = species) +
  geom_bar() +
  labs(x = "Species", y = "Count")
```

Besides species, the Palmer Penguins dataset also includes information about
the island where each bird was observed and the biological sex of each bird.
When a dataset has multiple categorical features, it's important to check how
the observations are grouped within them, since unbalanced groups can bias
analyses. You can use color to represent a second categorical feature in a bar
plot. In ggplot2, the fill color is set by the `fill` aesthetic. So to make the
bar plot show both species and sex:

```{r penguins-bar-species-sex}
ggplot(penguins) +
  aes(x = species, fill = sex) +
  geom_bar() +
  labs(x = "Species", y = "Count")
```

By default, the `geom_bar` function draws the bars stacked. If we want to
compare the groups across both species and sex, it's better to draw the bars
side-by-side. To do this, set `position = "dodge"` in the geometry:

```{r penguins-bar-species-sex-dodged}
ggplot(penguins) +
  aes(x = species, fill = sex) +
  geom_bar(position = "dodge") +
  labs(x = "Species", y = "Count")
```

The plot suggests the sexes are relatively well-balanced across the observed
penguins, regardless of species. Of course, there's still the island feature to
consider.

A bar plot can really only summarize two features at once, and many other types
of plots are also limited to just two or three features. One way to represent
more categorical features is to use **facets**, side-by-side plots that each
show data for a mutually exclusive category or combination of categories.

Most plotting packages provide convenience functions for making faceted plots,
and ggplot2 is no exception. There are two functions for facetting:

* The `facet_wrap` function displays a row of plots grouped by one categorical
  feature, wrapping to a new row whenever it runs out of horizontal space.
* The `facet_grid` function displays a grid of plots grouped by one or two
  categorical features arranged along the axes of the grid.

For example, let's use `facet_grid` to indicate sex by row, island by column,
and use a separate bar for each species:

```{r penguin-facet-species}
ggplot(penguins) +
  aes(x = species) +
  geom_bar(position = "dodge") +
  labs(x = "Species", y = "Count") +
  facet_grid(sex ~ island)
```

:::{.callout-tip}
If you want to change the names of the levels in a factor (a categorical
feature), do that as a separate step before making visualizations. DataLab's [R
Basics workshop reader][dl-r-basics] provides details about how to create
factors and set level names.

[dl-r-basics]: https://ucdavisdatalab.github.io/workshop_r_basics/
:::

ggplot2's functions for creating and customizing faceted plots are convenient,
and you can read more about them in the [documentation][ggplot2].

:::{.callout-note}
This case study focuses on categorical features and the next focuses on
continuous features. So what should you do if you have a discrete feature?

Discrete features share properties of categorical features (both are
enumerable) and properties of continuous features (both are quantitative). This
usually means you can choose whether to treat discrete features as categorical
or continuous. Categorical methods tend to be more appropriate for discrete
features that take relatively few distinct values, and continuous methods tend
to be more appropriate for ones that don't.
:::


## Case Study: Plotting a Continuous Distribution {#sec-plotting-continuous-distribution}

:::{.callout-note}
This case study shows how to:

* Plot the distribution of one or more continuous features
:::

@sec-plotting-categorical-distribution shows how to visualize the distribution
of a categorical feature. For this case study, let's examine one of the
continuous features in the Palmer Penguins dataset from
@sec-example-palmer-penguins-part-1. There are many different ways to visualize
continuous features, such as histograms, density plots, box plots, violin
plots, and empirical cumulative distribution function plots.

Let's visualize the distribution of the birds' body mass, which is recorded in
the `body_mass_g` feature. Some of the geometries for plotting continuous
distributions are:

* `geom_histogram` to make a histogram
* `geom_density` to make a density (or "kernel density estimator") plot
* `geom_boxplot` to make a box plot
* `stat_ecdf` (or `geom_step` with `stat = "ecdf"`) to make an empirical
  cumulative distribution function plot
* `geom_violin` to make a violin plot

For example, to make a histogram:

```{r penguin-hist}
ggplot(penguins) +
  aes(x = body_mass_g) +
  geom_histogram(color = "black") +
  labs(x = "Body Mass (g)")
```

The setting `color = "black"` in `geom_histogram` makes the ggplot2 draw black
borders around the histogram bars (the default is no borders).

To make a density plot instead, we can simply change the geometry:

```{r penguin-density}
ggplot(penguins) +
  aes(x = body_mass_g) +
  geom_density() +
  labs(x = "Body Mass (g)")
```

:::{.callout-tip}
If you want the bottom of the plot to be at y-axis 0, you can add a scale layer
with the `expand` parameter set to 0:

```{r penguin-density-scale-0}
ggplot(penguins) +
  aes(x = body_mass_g) +
  geom_density() +
  labs(x = "Body Mass (g)") +
  scale_y_continuous(expand = c(0, 0))
```

The `expand` parameter controls how much padding ggplot2 adds at the bottom and
top of the axis.
:::

Density plots are convenient when you want to break down the distribution of a
continuous feature across the categories of a categorical feature, because the
plot can show a separate line for each category. For instance, to group body
mass by species:

```{r penguin-density-by-species}
ggplot(penguins) +
  aes(x = body_mass_g, color = species) +
  geom_density() +
  labs(x = "Body Mass (g)", color = "Species")
```

This makes it easy to see that Gentoo penguins typically have more mass than
the other two species.

As in @sec-plotting-categorical-distribution, you can incorporate more
categorical features by faceting. To incorporate another continuous feature,
it's necessary to change plot types.

You can visualize the distributions of and relationship between two continuous
features with a scatter plot (`geom_point`) or a smoothed scatter plot (again
`geom_density_2d`). The latter smooths out the points in a scatter plot to show
their relative density, using a 2-dimensional generalization of the method used
to estimate a density plot. For instance, to plot body weight against flipper
length:

```{r penguin-density-2d}
ggplot(penguins) +
  aes(x = body_mass_g, y = flipper_length_mm, color = species) +
  geom_density_2d() +
  labs(
    x = "Body Mass (g)",
    y = "Flipper Length (mm)",
    color = "Species"
  )
```

In this plot it's possible to see how the distributions of body mass and
flipper length differ across species, as well as how body mass and flipper mass
are related.

:::{.callout-tip}
Visualizing more than two continuous features at once can be challenging. One
way to get around this problem is to create a separate plot for each pair of
features.

If it really is necessary to visualize more than two at once, consider
converting at least one of them into a categorical feature by **binning** or
**discretizing** the values. Then you can use strategies for categorical
features such as varying colors and faceting.
:::


## Case Study: Plotting a Time Series

:::{.callout-note}
This case study shows how to:

* Convert temporal data to appropriate data types
* Plot a time series as a histogram or a line
* Use multiple data sets in a single plot
* Customize and rotate the text of tick labels
:::

Plotting a time series can be tricky because you first need to make sure that
the temporal features have appropriate data types. For this case study, suppose
you want to make a visualization that shows the relationship between flu rates
in birds and humans.

People mail dead birds to the USDA and USGS, where scientists analyze the birds
to find out why they died. The USDA compiles the information into a public
[Avian Influenza data set][usda-birds] each year.

[usda-birds]: https://www.aphis.usda.gov/aphis/ourfocus/animalhealth/animal-disease-information/avian/avian-influenza/

:::{.callout-important}
You can download the Avian Influenza data [HERE][usda-birds-data].

[usda-birds-data]: https://ucdavis.box.com/s/babz1b5ak0kh7q7cktg7k5gzrt2ve9gc
:::

We'll use the readr package's `read_csv` function to read the data. Compared to
R's built-in `read.csv` function, this function is faster and more robust
against incorrect formatting.

The data's headers contain spaces, which are inconvenient for working in R.
With the `read_csv` function, we can set `name_repair = "universal"` to have it
automatically fix all column names so that they are syntactically valid:

```{r read-birds}
# install.packages("readr")
library("readr")

birds_path = "data/hpai-wild-birds-ver2.csv"
birds = read_csv(birds_path, name_repair = "universal")
str(birds)
```

Each row corresponds to one bird death. There are 8 columns with information
about the date, species of bird, collection method, and location. The
`Date.Detected` column is a date, while the rest of the columns are
categorical, which you can see in the structural summary.

The dates are in month-day-year order, so we can parse them with the lubridate
package's `mdy` function. We can use the dplyr package and R's built-in
`factor` function to concisely convert the remaining columns to factors:

```{r clean-birds, message=FALSE}
# install.packages("lubridate")
# install.packages("dplyr")
library("lubridate")
library("dplyr")

birds$Date.Detected = mdy(birds$Date.Detected)
birds = mutate(birds, across(-Date.Detected, factor))
```

:::{.callout-tip}
Specifying appropriate data types when you read data often provides performance
benefits, but doing data type conversions later gives you more flexibility to
clean the data first. So which approach you should use in any given situation
depends on the data set and your analysis goals.
:::

What kind of plot is appropriate for this dataset? Note that there are several
dates where more than one bird was collected:

```{r birds-date-table}
arrange(count(birds, Date.Detected), desc(n))
```

While you could use a line plot to show the counts for each date, there are
also many dates where the counts are zero. To see a smoothed out version of the
line plot, you can use a histogram of the observations:

```{r birds-histogram}
ggplot(birds) +
  aes(x = Date.Detected) +
  geom_histogram(color = "black")
```

The data set includes information about how the birds were collected in the
`Sampling.Method` column. You can incorporate this information in the plot by
setting the `fill` aesthetic. In this case, stacked bars are a good choice
since the focus is on the total counts for each date range rather than the
individual sampling methods. The code becomes:

```{r birds-histogram-stacked}
ggplot(birds) +
  aes(x = Date.Detected, fill = Sampling.Method) +
  geom_histogram(color = "black")
```

The plot looks good, so now let's add data about human flu rates. The CDC
collects data about flu hospitalizations across 13 states. The data is publicly
available as the [FluServ-NET data set][fluserv].

[fluserv]: https://gis.cdc.gov/GRASP/Fluview/FluHospRates.html

:::{.callout-important}
You can download the FluServ-NET data [HERE][fluserv-data].
:::

[fluserv-data]: https://ucdavis.box.com/s/a4m4ermwm0lqes2sh1jckdf16h0rpdkc

In the FluServ-NET data set, each row corresponds to a single combination of
week, year, age, sex, and race. There are 10 columns, which are a mix of data
types. The CSV file the CDC provides contains extra text at the beginning and
end, and the column names contain spaces and other problematic characters. You
can read and clean up the data set with this code:

```{r clean-flu}
# install.packages("stringr")
library("stringr")

flu_path = "data/FluSurveillance_Custom_Download_Data.csv"
flu = read_csv(flu_path, name_repair = "universal", skip = 2)

# Change column names to lowercase.
flu = rename_with(flu, str_to_lower)

# Remove the text at the end.
flu = filter(flu, catchment == "Entire Network")

# Convert the year and week columns to actual dates.
dates = make_date(year = flu$mmwr.year)
week(dates) = flu$mmwr.week
flu$date = dates
```

For the plot, we'll only need the `"Overall"` age, sex, and race categories:

```{r filter-flu}
flu = filter(flu,
  age.category == "Overall",
  sex.category == "Overall",
  race.category == "Overall"
)
```

Now we can add a line for the weekly flu rates in the `weekly_rate` column to
the bird mortality histogram. The weekly flu rates are measured in
hospitalizations per 100,000 people and typically range from 0 to 100.
Multiplying by 100 to convert to hospitalizations per 10 million people makes
the range 0 to 1000, which is a nice match for the y-axis already on the plot.

Since we're dealing with two datasets, we'll move the data and aesthetics
layers for each into the calls to the respective `geom_` functions. We'll also
add a title and a label to clarify what the y-axis means. The code becomes:

```{r example-flu-plotting}
ggplot() +
  geom_histogram(
    aes(x = Date.Detected, fill = Sampling.Method),
    birds,
    color = "black"
  ) +
  geom_line(aes(x = date, y = weekly.rate * 100), flu) +
  labs(
    x = "Date Detected",
    y = "Reported bird deaths\nHospitalizations per 10 million people"
  )
```


## Case Study: Plotting a Function

:::{.callout-note}
This case study shows how to:

* Plot a function
* Fill the area under a function
* Set custom colors for lines, fills, and annotations
* Hide the axes of a plot
* Add annotations to a plot
:::

Plotting a curve or function can make it much easier to understand and explain
its behavior. For this case study, suppose you want to make a visualization
that shows how the area of overlap between two different probability density
functions is a measure of how similar they are. They should have some overlap,
but not too much. The color palette should be UC blue (`#1a3f68`) and gold
(`#e6c257`) to match the rest of your presentation.

:::{.callout-note collapse="false"}
A *probability density function* is a function that shows how likely different
outcomes are for a continuous probability distribution. The probability of
outcomes in any given interval is the area under the curve. For example, if the
area under the curve between 0 and 1 is 0.2, then there's a 20% chance of an
outcome between 0 and 1. The total area under the curve is always 1.

Since the total area under a probability density function is always 1, the area
of overlap between two density functions is always between 0 and 1. As a
result, the area of overlap is a convenient measure of similarity.
:::

R provides probability density functions for common distributions. Let's use a
normal distribution and a gamma distribution. Normal distributions are widely
known, while gamma distributions are visually different but will still have
some overlap. The gamma distribution requires a shape parameter; let's use 10.

```{r initial-funcs}
f_norm = \(x) dnorm(x)
f_gamma = \(x) dgamma(x, 10)
```

We can use ggplot2's `geom_function` geometry to plot functions. This geometry
will automatically evaluate the function at many points over an interval. We'll
use 1,000 points in the interval $(-20, 20)$. We'll also set `linetype =
"dashed"` on one of the lines so that they're distinct. The code to make the
plot is:

```{r plot-initial-funcs}
blue = "#1a3f68"
gold = "#e6c257"
n = 1000
xlim = c(-20, 20)

ggplot() +
  geom_function(
    fun = f_norm,
    n = n, xlim = xlim, color = blue
  ) +
  geom_function(
    fun = f_gamma,
    n = n, xlim = xlim, color = blue,
    linetype = "dashed"
  )
```

The two distributions don't overlap much, but we can fix that by adjusting
their parameters. To get some overlap, let's increase the standard deviation
(`sd`) of the normal distribution to 4 and shift the location of the gamma
distribution left by 2. Then the code for the plot becomes:

```{r plot-funcs}
f_norm = \(x) dnorm(x, sd = 4)
f_gamma = \(x) dgamma(x + 2, 10)

blue = "#1a3f68"
gold = "#e6c257"
n = 1000
xlim = c(-20, 20)

ggplot() +
  geom_function(
    fun = f_norm,
    n = n, xlim = xlim, color = blue
  ) +
  geom_function(
    fun = f_gamma,
    n = n, xlim = xlim, color = blue,
    linetype = "dashed"
  )
```

Let's emphasize the overlap by filling it with a color. The area of overlap is
the area underneath whichever function happens to be smaller at a given
point---the minimum of the two functions. We can define a new function,
`f_min`, that gives the minimum. We'll use R's `pmin` function to compute the
element-wise minimum.

To add the `f_min` function to the plot with a fill color, we'll use the
`stat_function` statistic with `geom = "polygon"` instead of the
`geom_function` geometry. The layer for this function also needs to come first,
so that it's under the other layers. So the code becomes:

```{r plot-funcs-overlap}
f_norm = \(x) dnorm(x, sd = 4)
f_gamma = \(x) dgamma(x + 2, 10)
f_min = \(x) pmin(f_norm(x), f_gamma(x))

blue = "#1a3f68"
gold = "#e6c257"
n = 1000
xlim = c(-20, 20)

ggplot() +
  stat_function(
    fun = f_min,
    geom = "polygon",
    xlim = xlim, n = n, fill = gold
  ) +
  geom_function(
    fun = f_norm,
    xlim = xlim, n = n, color = blue
  ) +
  geom_function(
    fun = f_gamma,
    linetype = "dashed",
    xlim = xlim, n = n, color = blue
  )
```

This diagram is meant to make it easier to explain area of overlap, which is
shown by the curves and fill. The axis ticks and labels don't aid
understanding, so let's remove them. You can remove all decorations on a plot
by adding the `theme_void` theme. Let's also adjust the x interval to $(-15,
25)$ to center the curves:

```{r plot-funcs-overlap-nice}
f_norm = \(x) dnorm(x, sd = 4)
f_gamma = \(x) dgamma(x + 2, 10)
f_min = \(x) pmin(f_norm(x), f_gamma(x))

blue = "#1a3f68"
gold = "#e6c257"
n = 1000
xlim = c(-15, 25)

ggplot() +
  stat_function(
    fun = f_min,
    geom = "polygon",
    xlim = xlim, n = n, fill = gold
  ) +
  geom_function(
    fun = f_norm,
    xlim = xlim, n = n, color = blue
  ) +
  geom_function(
    fun = f_gamma,
    linetype = "dashed",
    xlim = xlim, n = n, color = blue
  ) +
  theme_void()
```

Finally, let's add a text label and arrow to emphasize the area of overlap. You
can use the `annotate` function to add an annotation to a plot. Annotations are
flexible: they can be points, lines, text, or almost any other geometry. With
the annotations, the plot becomes:

```{r plot-funcs-final}
f_norm = \(x) dnorm(x, sd = 4)
f_gamma = \(x) dgamma(x + 2, 10)
f_min = \(x) pmin(f_norm(x), f_gamma(x))

blue = "#1a3f68"
gold = "#e6c257"
n = 1000
xlim = c(-15, 25)

ggplot() +
  stat_function(
    fun = f_min,
    geom = "polygon",
    xlim = xlim, n = n, fill = gold
  ) +
  geom_function(
    fun = f_norm,
    xlim = xlim, n = n, color = blue
  ) +
  geom_function(
    fun = f_gamma,
    linetype = "dashed",
    xlim = xlim, n = n, color = blue
  ) +
  theme_void() +
  annotate(
    # Use the segment geometry to make a line.
    # Use the arrow parameter and arrow function to make an arrow.
    geom = "segment", arrow = arrow(ends = "first"),
    x = 5, xend = 15, y = 0.01, yend = 0.1,
    color = blue
  ) +
  annotate(
    geom = "text",
    # Multiply by 1.01 to nudge the position slightly.
    x = 15 * 1.01, y = 0.1 * 1.01,
    label = "overlap", size = 10,
    hjust = "left", vjust = "bottom",
    color = blue
  )
```


## Case Study: Error Bars

:::{.callout-note}
This case study shows how to:

* Add error bars to a plot
:::

Here's an example that recently came up in DataLab office hours. You've done an
experiment to see how mice with two different genotypes respond to two
different treatments. Now you want to plot the mean response of each group as a
column, with error bars indicating the standard deviation of the mean. You also
want to show the raw data. I've simulated some data for us to use---download it
[here][simulated-data].

[simulated-data]: https://ucdavis.box.com/s/lxgplwqn23ncbeb5gc6ktpfknr56xws6

This one is kind of complicated because you have to tell `ggplot2` how to
calculate the height of the columns and of the error bars.

```{r example-genotype, warning=FALSE, message=FALSE}
mice_geno = read_csv("data/genotype-response.csv")

# show the treatment response for different genotypes
ggplot(mice_geno) +
  aes(x=trt,
      y=resp,
      fill=genotype) +
  scale_fill_brewer(palette="Dark2") +
  geom_bar(position=position_dodge(),
           stat='summary',
           fun='mean') +
  geom_errorbar(fun.min=function(x) {mean(x) - sd(x) / sqrt(length(x))},
               fun.max=function(x) {mean(x) + sd(x) / sqrt(length(x))},
               stat='summary',
               position=position_dodge(0.9),
               width=0.2) +
  geom_point(position=
               position_jitterdodge(
                 dodge.width=0.9,
                 jitter.width=0.1)) +
  xlab("Treatment") +
  ylab("Response (mm/g)") +
  ggtitle("Mean growth response of mice by genotype and treatment")
```


## Case Study: Small Business Loans

:::{.callout-note}
This case study shows how to:

* Make a plot with a log-scale axis
:::

The US Small Business Administration (SBA) maintains data on the loans it
offers to businesses. Data about loans made since 2020 can be found at the
[Small Business Administration website][sba], or you can download it from
[here][sba-data]. We'll load that data and then explore some ways to visualize
it. Since the difference between a \$100 loan and a \$1000 loan is more like
the difference between \$100,000 and \$1M than between \$100,000 and \$100,900,
we should put the loan values on a  logarithmic scale. You can do this in
`ggplot2` with the `scale_y_log10` function (when the loan values are on the y
axis).

[sba]: https://data.sba.gov/dataset/7-a-504-foia
[sba-data]: https://ucdavis.box.com/s/ctx6grx10c83lh561ekl9tnmi20561ed


```{r example-SBA, message=FALSE, warning=FALSE}
# load the small business loan data
sba <- read_csv("data/small-business-loans.csv")

# check the SBA data to see the data types, etc.
head(sba)

# boxplot of loan sizes by business type
subset(sba, ProjectState == "CA") |>
  ggplot() +
  aes(x = BusinessType, y = SBAGuaranteedApproval) +
  geom_boxplot() +
  scale_y_log10() +
  ggtitle("Small Business Administraton guaranteed loans in California") +
  ylab("Loan guarantee (dollars)")


# relationship between loan size and interest rate
subset(sba, ProjectState == "CA") |>
  ggplot() +
  aes(x = GrossApproval, y = InitialInterestRate, ) +
  geom_point() +
  facet_wrap(~BusinessType, ncol = 3) +
  scale_x_log10() +
  ggtitle("Interest rate as a function of loan size") +
  xlab("Loan size (dollars)") +
  ylab("Interest rate (%)")
```


Now let's color the points by the loan status. Thankfully, `ggplot2` integrates
directly with [Color Brewer (colorbrewer2.org)](https://colorbrewer2.org) to
get better color palettes. We will use the `Accent` color palette, which is
just one of the many options that can be found on the Color Brewer site.

There are a lot of data points, which tent to largely overlap and hide each
other. We use a smoother (`geom_smooth`) to help call out differences that
would otherwise be lost in the noise of the points.

```{r example-SBA-facet, message=FALSE, warning=FALSE}
# color the dots by the loan status.
subset(sba, ProjectState == "CA" & LoanStatus != "EXEMPT" & LoanStatus != "CHGOFF") |>
  ggplot() +
  aes(x = GrossApproval, y = InitialInterestRate, color = LoanStatus) +
  geom_point() +
  geom_smooth() +
  facet_wrap(~BusinessType, ncol = 3) +
  scale_x_log10() +
  ggtitle("Interest rate as a function of loan size by loan status") +
  xlab("Loan size (dollars)") +
  ylab("Interest rate (%)") +
  labs(color = "Loan status") +
  scale_color_brewer(type = "qual", palette = "Accent")
```
